---
title: "0-Setup"
author: "Ethan Rogers"
date: "2025-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install R

This workflow was built with R version 4.5.1. To install R for your computer, follow the directions on the R Project [website](https://www.r-project.org).

## Clone this workflow to your preferred directory

This can be done by downloading the source zip file and unzipping the contents into your preferred directory.

## Run R

Navigate to the directory you added the workflow to above and launch R via command line. In bash or zsh, this can be done with the following commands, substituting `your/preferred/directory` with the directory path.

```
cd your/preferred/directory
R
```

## Install `renv`, create the virtual environment from the `renv.lock` file, and install dependencies

Now we can begin setting up our workspace and installing dependencies using `renv`. `renv` sets up isolated package libraries called virtual environments similar to python's `venv`. It's best practice to use a dedicated virtual environments for each project. However, in order to set up a virtual environment, we need to first install `renv` on the system library.

```{r, eval=FALSE}
install.packages("renv")
```

This workflow requires the following packages and their dependencies:

- Cardinal (installs its matter and BiocParallel backends)
- dplyr, tibble, tidyr, ggplot2, readr, stringr
- data.table
- matrixStats
- fuzzyjoin
- ggpubr, ggrepel, scales
- RColorBrewer, VennDiagram
- lmerTest, lme4, nlme
- emmeans
- broom, broom.mixed
- kableExtra
- knitr, rmarkdown

### Creating a `renv` and installing packages from the `renv.lock` file

`renv` will automatically detect the `renv.lock` file in the main directory, which will tell it to install the packages above and their dependencies. To commence this process, run the following:

```{r, eval=FALSE}
renv::init(bioconductor = "3.23")
```

### Creating a `renv` and letting `renv` detect and install used packages

**If the `renv.lock` is not present**, `renv` can automatically detect and install the required packages. In order to do this in `.Rmd` files `renv` needs `yaml`. **Keep in mind that this method does not preserve package versions, thus may not produce expected behavior.**

```{r, eval=FALSE}
install.packages("yaml")
renv::init(bioconductor = "3.23")
```

### Creating a `renv` and installing packages manually

Sometimes we need to install the packages ourselves manually. **Keep in mind that this method does not preserve package versions, thus may not produce expected behavior.**
```{r, eval = FALSE}
renv::init(bioconductor = "3.23", bare = T)
renv::activate() # or restart R

renv::install("bioc::Cardinal")

packages <- c("dplyr", "tibble", "tidyr", "ggplot2", "readr", "stringr",
              "data.table", "matrixStats", "fuzzyjoin",
              "ggpubr", "ggrepel", "scales",
              "RColorBrewer", "VennDiagram",
              "lmerTest", "lme4", "nlme",
              "emmeans", "broom", "broom.mixed",
              "kableExtra", "knitr", "rmarkdown")

renv::install(packages)
```


## Download data

The data are hosted in two places. The arthritis data for vignettes 1–5 is a reanalysis of MassIVE dataset [MSV000094448](https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c8ee2b4247994c4396007e1cffd0bcfb), accession [RMSV000000712](https://massive.ucsd.edu/ProteoSAFe/dataset_files.jsp?task=8d22a971a81c4655a50aacd262a6960e). The mouse stroke data for vignette 6 is hosted in a public Amazon S3 bucket. Both are downloaded directly, with no account or credentials.

Downloading is off by default, so knitting this document transfers nothing. To fetch data, set `download` to `TRUE` and, in each section below, set the items you want; the raw spectra of each dataset are selected by default. Intermediates are distributed as zip archives and expanded on download. This requires `curl` and either `unzip` or `tar`, all of which ship with current macOS, Windows, and Linux.

```{r}
download <- FALSE
```

### Create data directories

`Data/` holds the arthritis data for vignettes 1–5 and `Mouse-Stroke-Data/` the stroke data for vignette 6. Create the structure for both up front; the sections below download into it.

```{r}
data_dirs <- c(
  "Data/Raw", "Data/Preprocessing", "Data/Filtering-Aggregation", "Data/Simulation-Data",
  "Mouse-Stroke-Data/Raw", "Mouse-Stroke-Data/Preprocessing-Filtering"
)
for (d in data_dirs) dir.create(d, showWarnings = FALSE, recursive = TRUE)
```

### Download helpers

`fetch` downloads a single file, choosing FTPS or HTTPS from the URL, skipping files already present, and retrying transient failures. `fetch_par` downloads many files from MassIVE at once via `curl --parallel`, which is much faster than fetching them one at a time. `login` is passed to the MassIVE FTP server; `anonymous:anonymous` works for the public dataset, and can be set to the dataset's submission/reviewer credentials for a private one. `expand` unpacks a zip archive with whichever of `unzip` or `tar` is available, since R's internal unzip cannot read entries larger than 4 GB.

```{r}
fetch <- function(url, dest, login = "anonymous:anonymous", tries = 5) {
  if (file.exists(dest)) return(invisible())
  part <- paste0(dest, ".part")
  for (i in seq_len(tries)) {
    ok <- tryCatch({
      if (startsWith(url, "ftp")) {
        download.file(url, part, method = "curl",
                      extra = paste("--ssl-reqd --user", shQuote(login)),
                      mode = "wb", quiet = TRUE)
      } else {
        download.file(url, part, method = "libcurl", mode = "wb", quiet = TRUE)
      }
      file.exists(part)
    }, error = function(e) FALSE)
    if (isTRUE(ok)) break
    unlink(part)
    Sys.sleep(2 * i)
  }
  if (!file.exists(part)) stop("download failed: ", url)
  file.rename(part, dest)
}

fetch_par <- function(urls, dests, login = "anonymous:anonymous", max = 4, tries = 5) {
  keep <- !file.exists(dests)
  urls <- urls[keep]; dests <- dests[keep]
  if (!length(urls)) return(invisible())
  args <- c("--ssl-reqd", "--user", login, "--parallel", "--parallel-max", max,
            "--retry", tries, "--fail", "--silent", "--show-error", "--remove-on-error",
            c(rbind("-o", dests, urls)))
  if (system2("curl", args) != 0) stop("one or more downloads failed; rerun to resume")
}

expand <- function(zip, exdir = ".") {
  if (nzchar(Sys.which("unzip"))) {
    unzip(zip, exdir = exdir, unzip = "unzip")
  } else if (nzchar(Sys.which("tar"))) {
    system2("tar", c("-xf", shQuote(zip), "-C", shQuote(exdir)))
  } else {
    stop("Need 'unzip' or 'tar' to expand ", zip)
  }
}
```

### Arthritis data and Simulations (Vignettes 1–5 and Simulation 1-2)

Converted raw spectra and the intermediate result of each step. `Raw` and `Simulation-Data` are plain files; the others are zip archives that expand into `Data/Preprocessing` or `Data/Filtering-Aggregation`. Set an entry to `TRUE` to download it. The public dataset is downloaded anonymously — no account needed — and files are fetched in parallel, which is much faster than one at a time; anonymous access allows about `parallel_max = 4` simultaneous connections before the server refuses more. Still, this will take while.

```{r}
arthritis_url <- "ftp://massive-ftp.ucsd.edu/x01/RMSV000000712/2026-08-03_ebrogers_8d22a971"
massive_login <- Sys.getenv("MASSIVE_LOGIN", "anonymous:anonymous")
parallel_max  <- 4 # generally the max allowable by MassIVE

arthritis <- c(
  "Raw"                                                    = TRUE,
  "Individual-Picked-Aligned-Filtered"                     = FALSE,
  "Recalibrated-Picked-Aligned-FreqFiltered05perRun"       = FALSE,
  "normalized"                                             = FALSE,
  "Preprocessed-NSFiltered"                                = FALSE,
  "Preprocessed-NSFiltered-Aggregated"                     = FALSE,
  "Preprocessed-NSFiltered-Aggregated-CartilageFiltered50" = FALSE,
  "Simulation-Data"                                        = TRUE
)
```

```{r}
if (download) {
  samples <- c("6891a_lat_oa", "6891b_med_oa", "6908a_lat_oa", "6908b_med_oa",
               "6911a_lat_oa", "6911b_med_oa", "6919a_med_cad", "6919b_lat_cad",
               "6920a_lat_cad", "6920b_med_cad", "6924a_lat_cad", "6924b_med_cad",
               "6938a_med_cad", "6938b_lat_cad", "6944a_med_cad", "6944b_lat_cad",
               "6949a_lat_oa", "6949a_med_oa", "6954a_lat_oa", "6954b_med_oa")

  urls <- character(0); dests <- character(0)

  if (isTRUE(arthritis["Raw"])) {
    raw <- paste0(rep(samples, each = 2), c(".ibd", ".imzML"))
    urls  <- c(urls, paste0(arthritis_url, "/raw/", raw))
    dests <- c(dests, file.path("Data", "Raw", raw))
  }

  if (isTRUE(arthritis["Simulation-Data"])) {
    urls  <- c(urls, paste0(arthritis_url, "/metadata/sim_list_original.RData"))
    dests <- c(dests, file.path("Data", "Simulation-Data", "sim_list_original.RData"))
  }

  zips <- setdiff(names(which(arthritis)), c("Raw", "Simulation-Data"))
  ztmp <- file.path(tempdir(), paste0(zips, ".zip"))
  urls  <- c(urls, paste0(arthritis_url, "/raw/", zips, ".zip"))
  dests <- c(dests, ztmp)

  fetch_par(urls, dests, massive_login, parallel_max)

  for (t in ztmp) { expand(t); unlink(t) }
}
```

### Mouse stroke data (vignette 6)

Converted raw spectra and intermediates for the stroke case study, all zip archives that expand into `Mouse-Stroke-Data`. Set an entry to `TRUE` to download it.

```{r}
stroke_url <- "public URL available soon"

stroke <- c(
  "Raw_imzML"              = TRUE,
  "Recalibrated"           = FALSE,
  "Aligned"                = FALSE,
  "Analytes-Aligned"       = FALSE,
  "IS-Aligned"             = FALSE,
  "Analytes-IS-Aligned"    = FALSE,
  "Aligned-Tissue-ROIs"    = FALSE,
  "IS-Aligned-Tissue-ROIS" = FALSE,
  "IS-normalized"          = FALSE,
  "IS-normalized-filtered" = FALSE
)
```

```{r}
if (download) {
  for (z in names(which(stroke))) {
    tmp <- file.path(tempdir(), paste0(z, ".zip"))
    fetch(paste0(stroke_url, "/", z, ".zip"), tmp)
    expand(tmp)
    unlink(tmp)
  }
}
```
