---
title: "0-Setup"
author: "Ethan Rogers"
date: "2025-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install R

This workflow was built with R version 4.5.1. To install R for your computer, follow the directions on the R Project [website](https://www.r-project.org). 

## Clone this workflow to your preferred directory

This can be done by downloading the source zip file and unzipping the contents into your preferred directory.

## Run R

Navigate to the directory you added the workflow to above and launch R via command line. In bash or zsh, this can be done with the following commands, substituting `your/preferred/directory` with the directory path.

```
cd your/preferred/directory
R
```

## Install `renv`, create the virtual environment from the `renv.lock` file, and install dependencies

Now we can begin setting up our workspace and installing dependencies using `renv`. `renv` sets up isolated package libraries called virtual environments similar to python's `venv`. It's best practice to use a dedicated virtual environments for each project. However, in order to set up a virtual environment, we need to first install `renv` on the system library.

```
install.packages("renv")
renv::init(bioconductor = TRUE)
```

This workflow requires the following packages and their dependencies:
- Cardinal
- dplyr, tibble, tidyr, and ggplot2
- data.table
- fuzzyjoin
- patchwork, ggpubr, ggrepel
- lmerTest, nlme
- emmeans

`renv` will automatically detect the `renv.lock` file in the main directory, which will tell it to install the packages above and their dependencies. To commence this process, run the following:

## Download data from MassIVE

In order to run this workflow, we need to download the dataset raw .imzML files. This workflow is written to look for data in the `./Data` directory in the working directory.

You can create it manually, or run the following R code. If they exist already, they will not be overwritten. 

```{r}
dirs <- c(
  file.path(".", "Data"),
  file.path(".", "Data", "Raw"),
  file.path(".", "Data", "Filtering-Aggregation"),
  file.path(".", "Data", "Preprocessing"),
  file.path(".", "Data", "Simulation-Data"),
  file.path(".", "Outputs"),
  file.path(".", "Recreated-Figures")
)

for (d in dirs) {
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
}
```


The original dataset in Bruker .d format can be found on MassIVE via ID number: [MSV000094448](https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c8ee2b4247994c4396007e1cffd0bcfb). The reanalysis container holds all intermediates, including raw .imzML images that will be used for this analysis.

You can download all of them, or specific files:

```{r}
ftp_url <- "ftp://massive-ftp.ucsd.edu/v07/MSV000094448/"

file_list <- readLines(url(ftp_url))

# Parse filenames
filenames <- trimws(sub(".*\\s+", "", file_list))
filenames <- filenames[filenames != "" & !filenames %in% c(".", "..")]

# make destination dir if it doesn't exist
dest_dir <- file.path(".", "Data")
dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)

# Download each file
for (f in filenames) {
  remote <- paste0(ftp_url, f)
  local  <- file.path(dest_dir, f)
  temp   <- paste0(local, ".part")
  
  # Skip if the final file already exists
  if (file.exists(local)) {
    message("Already exists, skipping: ", f)
    next
  }
  
  # Remove any partial download from a previous interrupted attempt
  if (file.exists(temp)) {
    file.remove(temp)
  }
  
  message("Downloading: ", f)
  tryCatch({
    download.file(remote, destfile = temp, mode = "wb", method = "libcurl")
    # Only rename to final name if download completed successfully
    file.rename(temp, local)
  }, error = function(e) {
    message("Failed to download: ", f, " - ", conditionMessage(e))
    # Clean up partial file
    if (file.exists(temp)) file.remove(temp)
  })
}
```