---
title: "0-Setup"
author: "Ethan Rogers"
date: "2025-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install R

This workflow was built with R version 4.5.1. To install R for your computer, follow the directions on the R Project [website](https://www.r-project.org). 

## Clone this workflow to your preferred directory

This can be done by downloading the source zip file and unzipping the contents into your preferred directory.

## Run R

Navigate to the directory you added the workflow to above and launch R via command line. In bash or zsh, this can be done with the following commands, substituting `your/preferred/directory` with the directory path.

```
cd your/preferred/directory
R
```

## Install `renv`, create the virtual environment from the `renv.lock` file, and install dependencies

Now we can begin setting up our workspace and installing dependencies using `renv`. `renv` sets up isolated package libraries called virtual environments similar to python's `venv`. It's best practice to use a dedicated virtual environments for each project. However, in order to set up a virtual environment, we need to first install `renv` on the system library.

```
install.packages("renv")
renv::init(bioconductor = TRUE)
```

This workflow requires the following packages and their dependencies:

- Cardinal
- dplyr, tibble, tidyr, and ggplot2
- data.table
- fuzzyjoin
- patchwork, ggpubr, ggrepel
- lmerTest, nlme
- emmeans

`renv` will automatically detect the `renv.lock` file in the main directory, which will tell it to install the packages above and their dependencies. To commence this process, run the following:

## Download data from S3
The original dataset in Bruker .d format can be found on MassIVE via ID number: [MSV000094448](https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c8ee2b4247994c4396007e1cffd0bcfb). Due to some technical issues with MassIVE and .imzML files, we have stored the converted raw files and all intermediates created from this analysis in an public Amazon S3 bucket. Files can be downloaded directly over HTTPS — no AWS account, credentials, or additional software is required.

This workflow is written to look for data in the `./Data` directory, with specific steps having sub directories. If these already exist this code will do nothing, otherwise we will create them.

### Create local directories
```{r}
dirs <- c(
  file.path(".", "Data"),
  file.path(".", "Data", "Raw"),
  file.path(".", "Data", "Filtering-Aggregation"),
  file.path(".", "Data", "Preprocessing"),
  file.path(".", "Data", "Simulation-Data"),
  file.path(".", "Outputs"),
  file.path(".", "Recreated-Figures")
)

for (d in dirs) {
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
}
```

### Select which data to download

You do not need to download everything. Set any of the following to `FALSE` to skip that directory. At minimum, you will need `Raw` and `Simulation-Data` to run the full workflow from scratch, or the later intermediates to reproduce specific steps. T

```{r}
download_dirs <- list(
  "Data/Raw"                        = TRUE,
  "Data/Preprocessing/Smoothed-BLR-Recalibrated" = TRUE,
  "Data/Preprocessing/Individual-Picked-Aligned-Filtered" = TRUE,
  "Data/Preprocessing/Smoothed-BLR-Recalibrated-Picked-Aligned-FreqFiltered05perRun.imzML" = TRUE,
  "Data/Preprocessing/normalized.imzML" = TRUE,
  "Data/Filtering-Aggregation/Preprocessed-NSFiltered.imzML" = TRUE,
  "Data/Filtering-Aggregation/Preprocessed-NSFiltered-Aggregated.imzML" = TRUE,
  "Data/Filtering-Aggregation/Preprocessed-NSFiltered-Aggregated-CartilageFiltered25.imzML" = TRUE,
  "Data/Simulation-Data"            = TRUE
)
```

### Download helper functions

The following functions list and download files from the public S3 bucket using only base R. No additional packages are needed.

```{r}
bucket    <- "preprint-msi-arthtitis-data"
base_url  <- paste0("https://", bucket, ".s3.amazonaws.com")
source("S3Utilities.R")
```

### Run the download
The following code iterates over your selected directories, lists the files in each from the S3 bucket, and downloads them. Change `dry_run` to `FALSE` to download the files selected from above.

```{r}
dry_run <- TRUE

selected <- names(download_dirs)[unlist(download_dirs)]

for (prefix in selected) {
  message("\n--- Listing: ", prefix, " ---")
  query_prefix <- ifelse(grepl("/$", prefix), prefix, paste0(prefix, "/"))
  objects <- s3_list_objects(base_url, query_prefix)

  if (nrow(objects) == 0) {
    message("  No files found for prefix: ", prefix)
    next
  }

  total_size <- sum(objects$size)
  already_exist <- vapply(objects$key, function(k) file.exists(file.path(".", k)), logical(1))
  n_skip <- sum(already_exist)
  n_todo <- nrow(objects) - n_skip

  message(sprintf("  %d files (%s total), %d to download, %d already present",
                  nrow(objects), human_size(total_size), n_todo, n_skip))

  for (i in seq_len(nrow(objects))) {
    key <- objects$key[i]
    size <- objects$size[i]
    if (file.exists(file.path(".", key))) next

    if (dry_run) {
      message(sprintf("  [DRY RUN %d/%d] would download: %s (%s)",
                      i, nrow(objects), basename(key), human_size(size)))
    } else {
      s3_download_file(base_url, key, dest_dir = ".")
    }
  }
}
```
