---
title: "5-Designing-Future-Experiments"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

Designing future experiments requires translating what we learned from the current data into practical planning quantities. Here, I use the fitted models from the cartilage analysis to estimate how many biological replicates a future study would need to detect differences of various magnitudes. The goal is not to produce a single universal sample size, but to show how the required number of subjects changes as the anticipated effect size and multiple-testing burden change.

# Load required packages

```{r library, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(Cardinal)
library(lmerTest)
library(emmeans)
source("Utilities.R")
```

# Load model results and restrict to adequately fit features

Power calculations are only as reasonable as the variance estimates that feed them. For that reason, I begin with the fitted cartilage models and retain only features that passed the model checks from the previous chapter. This keeps poorly behaved models from disproportionately influencing the planning values.

```{r load-results}
load(file.path("Outputs", "cartilage_results.RData"))

signift_cartilage.RData

model_fit_eval <- read_csv("Outputs/feature-model-fit-evaluation.csv", show_col_types = FALSE) |> 
  mutate(across(qq_normality:outliers, as.factor))

passed <- model_fit_eval |> 
  filter(qq_normality == "OK" & 
           residual_variance == "OK" & 
           outliers == "OK")

cartilage_models_OK_fit <- cartilage_res[[1]][passed$feature_id]
```

# Extract planning quantities from the fitted models

The previous mixed models separate variation attributable to subjects from residual variation. That distinction matters here because the two future comparisons have different variance structures:

- **In Medial: Osteoarthritis vs Control** is a between-subject comparison, so both subject-level and residual variation contribute.
- **In Osteoarthritis: Medial vs Lateral** is a within-subject comparison, so the design is less affected by subject-to-subject heterogeneity.

To obtain stable planning values, we can summarize the variance components across adequately fit features and use their medians rather than selecting any single feature. I also compute a typical reference mean estimate so that differences on the intensity scale can be expressed as percent differences, which are easier to interpret.

```{r extract-variance-components}
df_var <- bind_rows(lapply(cartilage_models_OK_fit, function(m) {
  vc <- as.data.frame(VarCorr(m))
  sigma2_subject <- vc$vcov[1]
  sigma2_residual <- attr(VarCorr(m), "sc")^2
  tibble(sigma2_residual = sigma2_residual,
         sigma2_subject = sigma2_subject,
         intercept = coef(summary(m))[1, 1],
         condition = abs(coef(summary(m))[2, 1]),
         subtissue = abs(coef(summary(m))[3, 1]))
})) %>% mutate(id = "Cartilage", feature_id = row_number())

FDR <- 0.05
power <- 0.9

median.sigma.subject <- median(df_var$sigma2_subject, na.rm = TRUE)
median.sigma.error <- median(df_var$sigma2_residual, na.rm = TRUE)
median_denom_for_pct_delta <- median(signift_cartilage$reference_estimate, na.rm = TRUE)
```

# Define the sample size calculation

I will use a simple normal-approximation calculation to estimate the number of biological replicates needed to detect an anticipated difference with 90% power. The function below evaluates a range of possible effect sizes so that we can visualize the tradeoff between detectable percent difference and required sample size.

```{r define-sample-size-function}
calc_sample_size <- function(desired_difference,
                             alpha,
                             power,
                             variance_estimate_without_I) {
  delta <- seq(desired_difference[1], desired_difference[2], 0.1)
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(power)
  aa <- z_alpha + z_beta
  denom <- (delta / aa)^2
  numSample <- variance_estimate_without_I / denom
  data.frame(delta, numSample, FDR, power)
}
```

# Specify multiple-testing scenarios and target comparisons

In practice, future studies will rarely test just one feature. The effective significance threshold therefore depends on how severe the multiple-testing burden is expected to be. I consider three scenarios:

- **No correction, i.e. a single comparison**, which serves as a best-case benchmark.
- **FDR correction, 50% true negative rate**, representing a moderate multiple-testing burden.
- **FDR correction, 90% true negative rate**, representing a more stringent setting in which most tested features are expected not to change.

These scenarios are encoded through different alpha values. I then evaluate them for the two comparisons of interest from the previous chapters.

```{r define-scenarios}
alpha_1 <- power * FDR / (1 + (1 - FDR) * 1)
alpha_90 <- power * FDR / (1 + (1 - FDR) * 90)
alpha_50 <- power * FDR / (1 + (1 - FDR) * 50)

comparison_labels <- c("In Osteoarthritis: Medial vs Lateral",
                       "In Medial: Osteoarthritis vs Control")
```

The between-subject comparison uses `2*(median.sigma.error + median.sigma.subject)`, whereas the within-subject comparison uses `2*(median.sigma.error)`. This reflects the fact that pairing within subject removes some of the variation that would otherwise inflate the required sample size.

```{r build-power-curves}
df_curves <- bind_rows(
  calc_sample_size(c(1, 150),
                   alpha_90,
                   power,
                   2 * (median.sigma.error + median.sigma.subject)) %>%
    mutate(id = "Condition", n_sig_test = 90),

  calc_sample_size(c(1, 150),
                   alpha_90,
                   power,
                   2 * (median.sigma.error)) %>%
    mutate(id = "Subsection", n_sig_test = 90),

  calc_sample_size(c(1, 150),
                   alpha_50,
                   power,
                   2 * (median.sigma.error + median.sigma.subject)) %>%
    mutate(id = "Condition", n_sig_test = 50),

  calc_sample_size(c(1, 150),
                   alpha_50,
                   power,
                   2 * (median.sigma.error)) %>%
    mutate(id = "Subsection", n_sig_test = 50),

  calc_sample_size(c(1, 150),
                   alpha_1,
                   power,
                   2 * (median.sigma.error + median.sigma.subject)) %>%
    mutate(id = "Condition", n_sig_test = 1),

  calc_sample_size(c(1, 150),
                   alpha_1,
                   power,
                   2 * (median.sigma.error)) %>%
    mutate(id = "Subsection", n_sig_test = 1)
) %>%
  mutate(comparison = factor(ifelse(id == "Subsection",
                                    comparison_labels[1],
                                    comparison_labels[2]),
                             levels = comparison_labels),
         n_sig_test = factor(n_sig_test,
                             levels = c(1, 50, 90),
                             labels = c("No correction, i.e. a single comparison",
                                        "FDR correction, 50% true negative rate",
                                        "FDR correction, 90% true negative rate")))
```

# Visualize how effect size and multiple testing affect sample size

To make the curves easier to interpret, I convert the anticipated absolute intensity difference into a percent difference relative to a typical reference mean. The x-axis therefore represents the absolute value of the anticipated percent difference in mean intensity, and the y-axis gives the minimum number of biological replicates required for 90% power. The horizontal line at 4 marks a practically attractive target, since small MSI studies are often constrained by cost, throughput, or tissue availability.

The line types show how the required sample size changes as the multiple-testing burden becomes more stringent. The two colors distinguish the within-subject and between-subject comparisons. If the darker curve lies above the lighter one, that indicates the between-subject comparison is harder to power and requires more subjects for the same anticipated effect size.

```{r power-curve-plot}
greys <- c("In Osteoarthritis: Medial vs Lateral" = "grey70",
           "In Medial: Osteoarthritis vs Control" = "grey30")

ggplot(df_curves,
       aes(x = delta / median_denom_for_pct_delta,
           y = numSample,
           color = comparison)) +
  geom_line(aes(linetype = n_sig_test), linewidth = 0.9) +
  geom_hline(yintercept = 4, alpha = 0.5, linewidth = 0.7) +
  scale_linetype_manual(values = c("solid", "31", "11"),
                        name = "Multiple Testing Correction") +
  theme_bw() +
  ylab("Minimum Number of Biological Replicates") +
  xlab("Absolute Value of Anticipated Percent Difference of Mean Intensities") +
  ylim(0, 20) +
  labs(color = "Comparison") +
  theme(legend.position = "bottom") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0.11, 0.5)) +
  scale_color_manual(values = greys) +
  guides(color = guide_legend(order = 1, nrow = 2,
                              theme = theme(legend.title.position = "top")),
         linetype = guide_legend(order = 2, nrow = 3,
                                 theme = theme(legend.title.position = "top")))

ggsave(file.path("Recreated-Figures",
                 "Figure-Designing-Future-Experiments-Power.png"),
       height = 6,
       width = 6,
       units = "in")
```

As expected, increasing the multiple-testing burden shifts the curves upward, meaning that stricter control of false discoveries requires more subjects to detect the same effect size. Likewise, the between-subject comparison is generally more demanding than the within-subject comparison because it carries both residual and subject-level variability.

# Identify effect sizes that correspond to roughly 4 biological replicates

One practical way to read the figure is to ask what size difference would be needed to keep the study near 4 biological replicates under the more stringent correction setting. The table below extracts the point on each 90% true-negative-rate curve nearest that target.

```{r summarize-four-replicate-threshold}
df_curves %>% 
  filter(n_sig_test == "FDR correction, 90% true negative rate") %>% 
  group_by(comparison) %>% 
  filter(abs(numSample - 4) == min(abs(numSample - 4))) %>% 
  mutate(pct_delta = delta / median_denom_for_pct_delta)
```

These thresholds provide a useful planning heuristic. If a future study expects effects at least this large, then a design with roughly 4 biological replicates may be realistic even after accounting for multiple testing. If the expected effects are smaller, then the curves indicate that more replicates will likely be necessary.

# Take-home interpretation

The main lesson from this exercise is that study design depends strongly on both the expected effect size and the testing burden. Within-subject contrasts are easier to power than between-subject contrasts, and aggressive multiple-testing correction can quickly raise sample size requirements. For future MSI experiments, these curves are best used as planning guides: choose the comparison of interest, decide what effect size is scientifically meaningful, and then read off the approximate number of biological replicates required under a realistic multiple-testing scenario.