---
title: "6-Mouse-Stroke-Case-Study"
author: "Ethan Rogers"
date: "06/29/2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

In this vignette we demonstrate the proposed workflow on a case study of mouse brain samples after photothrombotic stroke.

# Step 0: Setup

Please follow the directions in vignette `0-Setup.Rmd` to create the environment and directory trees.

# Step 1: Preprocessing

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(data.table)
library(Cardinal)

source("Utilities.R")
source("SpatialUtilities.R")

# from wd
data.path <- file.path("Mouse-Stroke-Data")
raw.data.path <- file.path(data.path ,"Raw")
preprocessing.data.path <- file.path(data.path ,"Preprocessing")
recalib.data.path <- file.path(preprocessing.data.path,"Recalibrated")
aligned.data.path <- file.path(preprocessing.data.path,"Aligned")

outputs.path <- file.path("Outputs")
```

## Load metadata
```{r}
name_tibble <- read_csv(file.path(outputs.path, "mouse_stroke_metadata.csv"), show_col_types = FALSE)
```

## Loading data

```{r}
samp <- name_tibble |>
    filter(!is.na(days) & !is.na(replicate))
```

```{r}
msa.raw <- list()

for (s in samp$filename) {
    m <- readMSIData(file.path(raw.data.path, s))

    # reorient brains to be vertical
    crd <- coord(m)
    if (max(crd$x) - min(crd$x) < max(crd$y) - min(crd$y)) {
        coord(m)$x <- crd$y
        coord(m)$y <- crd$x
    }

    # rename run to be sample short name
    sn <- name_tibble |> 
        filter(filename == s) |>
        pull(short_name)

    runNames(m) <- sn
    msa.raw[[sn]] <- m
}

gc()
```

## Recalibration
These data do not require baseline removal or smoothing.

```{r, eval=FALSE}
bp <- SnowfastParam(workers = 10L, force.GC = T)

for (sn in samp$short_name[3:10]) {
    print(paste("Starting", sn, "\n"))
    est_ref_peaks <- msa.raw[[sn]] |>
		estimateReferencePeaks(BPPARAM = bp)
    msa.raw[[sn]] |>
	    recalibrate(ref = est_ref_peaks, tolerance = 10, units = "ppm") |>
        process(BPPARAM = bp) |>
        writeMSIDataSafe(file.path(recalib.data.path, paste0(sn, ".imzML")))
}

gc()
```

# Peak alignment from known peak list and IS C13 lipids

```{r}
lcms.peaks <- read_csv(file.path(outputs.path, 
                    "LCMSlipids_brain_wang_et_al_nature_metab_2025.csv"),
                    show_col_types = FALSE)

#for normalization
c13.is.lipids <- c(428.299845, 499.355465, 596.372775, 708.589835, 777.629855, 785.6367, 800.6476, 876.662865)

all.peaks.to.align <- sort(c(c13.is.lipids, lcms.peaks$`Input Mass`))
```

```{r, eval=FALSE}
bp <- SnowfastParam(workers = 8L, force.GC = T)

# load, align, save
for (sn in samp$short_name) {
    readMSIData(file.path(recalib.data.path, paste0(sn, ".imzML"))) |>
    peakAlign(ref = all.peaks.to.align) |>
    process(BPPARAM = bp) |>
    writeMSIDataSafe(file.path(aligned.data.path, paste0(sn, ".imzML")))
}
gc()
```

```{r}
ind.c13.is.lipids <- features(readMSIData(file.path(aligned.data.path, paste0(samp$short_name[1], ".imzML"))),
                                mz = c13.is.lipids)
```

```{r}
# keep the original 146 lipids

mse_list.all <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML")))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse.all <- Reduce(cbind, mse_list.all)

writeMSIDataSafe(mse.all, file.path(preprocessing.data.path, "Analytes-IS-Aligned.imzML"))
```

```{r}
# keep the original 146 lipids

mse_list <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML"))) |>
    subsetFeatures(!(1:length(all.peaks.to.align) %in% ind.c13.is.lipids)) |>
    summarizePixels(stat = c("nnzero", "mean", "sd"), na.rm = T, chunkopts = list(nchunks = 1))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse <- Reduce(cbind, mse_list)

writeMSIDataSafe(mse, file.path(preprocessing.data.path, "Analytes-Aligned.imzML"))
```

```{r}
# keep the 8 IS lipids

mse_list.is <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML"))) |>
    subsetFeatures(1:length(all.peaks.to.align) %in% ind.c13.is.lipids) |>
    summarizePixels(stat = c("nnzero", "mean", "sd"), na.rm = T, chunkopts = list(nchunks = 1))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse.is <- Reduce(cbind, mse_list.is)

writeMSIDataSafe(mse.is, file.path(preprocessing.data.path, "IS-Aligned.imzML"))
```


## ROI selection

```{r}
key <- function(d) paste(d$run, d$x, d$y, sep = "\r")
```

### Trim background 

```{r}
intensity(mse) <- as.matrix(intensity(mse))

mse <- mse |> add_edge_factor_mse(radius = 10,  contiguous = FALSE)

mse_subset <- mse |> 
                subsetPixels(((edge == F & nnzero > 78) | (edge == T & nnzero > 88)))

islands_drop_mask <- drop_pixel_islands_mask_mse(mse_subset, n = 20)

mse_subset_trim_islands <- mse_subset |> subsetPixels(islands_drop_mask)

tissue_mask <- key(pData(mse)) %in% key(pData(mse_subset_trim_islands))
```

```{r}
#for (i in 1:146) {
#    image_png(mse_subset_trim_islands, i = i, style = "dark", bg = "black", key = F, grid = F, layout = c(5, 2), filename = paste("tmp/feature_", i, ".png"),
#          width = 1500, height = 2200, pointsize = 24)
#}
```

### Merge annotations from QuPath

We used 834.5273 m/z (PS 40:6) to segment stroke from non-stroke area as it was previously [identified as a marker of white matter](https://doi.org/10.1038/s41598-018-37612-5) in a mouse stroke model.

```{r}
annot.qp <- read_csv(file.path(outputs.path, 
                    "stroke-hemisphere-cortexwm-annotations-qupath.csv"),
                    show_col_types = FALSE)
```

```{r}
# verify order (all should be true)
all(pData(mse_subset_trim_islands)$x == annot.qp$x)
all(pData(mse_subset_trim_islands)$y == annot.qp$y)
all(pData(mse_subset_trim_islands)$run == annot.qp$run)

pData(mse_subset_trim_islands)$stroke <- annot.qp$stroke
pData(mse_subset_trim_islands)$hemisphere <- annot.qp$hemisphere
pData(mse_subset_trim_islands)$cortexwm <- annot.qp$cortexwm
```

```{r}
image(mse_subset_trim_islands, "stroke")
image(mse_subset_trim_islands, "hemisphere")
image(mse_subset_trim_islands, "cortexwm")
```

### Trim areas with very high internal standard lipid after tissue mask

```{r}
intensity(mse.is) <- as.matrix(intensity(mse.is))

mse.is.tmp <- mse.is |>
    subsetPixels(tissue_mask) |>
    summarizeFeatures(stat = c("nnzero", "sd", "mean"), na.rm = T) |>
    subsetFeatures(mean > 200) |>
    summarizePixels(stat = c("nnzero", "sd", "mean"), na.rm = T)

hist(mse.is.tmp$mean)

quantile(mse.is.tmp$mean, p = .99)

is.mask <- mse.is.tmp$mean < quantile(mse.is.tmp$mean, p = .99)
```

```{r, eval=FALSE}
image_png(mse.is |> subsetPixels(tissue_mask) , subset = is.mask, i = c(1,4,5,7,8), layout = c(5, 10), key = F, grid = F, style = "dark", bg = "black", filename = "is.lipids.ltp99.png", width = 10000, height = 6000, pointsize = 24)
```

```{r}
mse.roi.trimmed <- mse_subset_trim_islands |> 
    subsetPixels(is.mask)

image(mse.roi.trimmed, "stroke")
image(mse.roi.trimmed, "hemisphere")

writeMSIDataSafe(mse.roi.trimmed, file.path(preprocessing.data.path, "Aligned-Tissue-ROIs.imzML"))
```


```{r}
mse.is.roi.trimmed <- mse.is |>
    subsetPixels(tissue_mask) |>
    subsetPixels(is.mask)

writeMSIDataSafe(mse.is.roi.trimmed , file.path(preprocessing.data.path, "IS-Aligned-Tissue-ROIS.imzML"))
```

```{r}
#final checks
all(mse.roi.trimmed$run == mse.is.roi.trimmed$run)
all(mse.roi.trimmed$x == mse.is.roi.trimmed$x)
all(mse.roi.trimmed$y == mse.is.roi.trimmed$y)
```


## Normalization


```{r}
#keeping these high abundant IS lipids
c13.is.lipids[c(1,4,5,7,8)]

mse.all <- readMSIData(file.path(preprocessing.data.path, "Analytes-IS-Aligned.imzML"))

intensity(mse.all) <- as.matrix(intensity(mse.all))

mask <- key(pData(mse.all)) %in% key(pData(mse.roi.trimmed))

mse.all.trimmed <- mse.all |>
    subsetPixels(mask)

mse.all.trimmed$stroke <- mse.roi.trimmed$stroke
mse.all.trimmed$hemisphere <- mse.roi.trimmed$hemisphere
mse.all.trimmed$cortexwm <- mse.roi.trimmed$cortexwm

# label features
n.features <- nrow(fData(mse.all.trimmed))

fData(mse.all.trimmed)$isIS <- 1:n.features %in% features(mse.all.trimmed, mz=c13.is.lipids)

fData(mse.all.trimmed)$isMarker <- 1:n.features %in% features(mse.all.trimmed, mz=c(834.5273))
```

```{r}
normalizeMultipleRefs <- function(object, refs, scale = 1, ...) {
    if ( missing(refs) || is.null(refs) || length(refs) == 0L )
        stop("must provide one or more reference values via 'refs'")
    refs <- sort(unique(as.numeric(refs)))
    FUN <- function(x, t, refs, scale = 1, ...) {
        if ( any(refs < min(t)) || any(refs > max(t)) )
            stop("one or more 'refs' is outside of domain limits")
        i <- matter::bsearch(refs, t, nearest = TRUE)
        b <- mean(x[i], na.rm = TRUE)
        if ( !is.na(b) && b != 0 ) {
            y <- scale * x / b
        } else {
            y <- rep.int(0, length(x))
        }
        dim(y) <- dim(x)
        y
    }
    Cardinal::addProcessing(object, FUN,
        label = "intensity normalization",
        metadata = list(method = "multiref", refs = refs),
        refs = refs, scale = scale, ...)
}
```

```{r}
#drop features before normalization
prenormalized <- mse.all.trimmed |>
    summarizeFeatures() |>
    subsetFeatures(!isMarker & !(isIS & mean < 200)) 

normalized <- prenormalized |>
    normalizeMultipleRefs(mz(prenormalized)[fData(prenormalized)$isIS], scale = 1000) |>
    process()

```

```{r, eval=FALSE}
for (i in 1:length(mz(normalized))) {
    image_png(normalized, i = i, style = "dark", bg = "black", key = F, grid = F, layout = c(5, 2), filename = paste("tmp/normalized/feature_", i, ".png"),
          width = 1500, height = 2200, pointsize = 24)
}
```

```{r}
writeMSIDataSafe(normalized, file.path(preprocessing.data.path, "IS-normalized.imzML"))
```

```{r}
knitr::knit_exit()
```

# ANALYSIS: between stroke day conditions

# Step 2: Filtering and aggregation
## Non-specific Filtering

```{r}
normalized <- readMSIData(file.path(preprocessing.data.path, "IS-normalized.imzML"))

intensity(normalized) = as.matrix(intensity(normalized))
```

Filter marker features and IS
```{r}
normalized <- normalized |>
            subsetFeatures(!isMarker & !isIS)
```

### Identify low mean and low sd features
Features in the 10th percentile or less for both mean and sd.

```{r}
normalized <- normalized |>
    summarizeFeatures(stat = c("sd", "nnzero", "mean"), na.rm = T, chunkopts = list(nchunks = 1))

mean.sd.p1 <- normalized |>
    fData() |>
    as_tibble() |>
    mutate(fid = row_number()) |>
    filter(mean < quantile(mean, p = .1) & sd < quantile(sd, p = .1)) |>
    pull(fid) |>
    unique()
```

### Identify features with sparse stroke and/or cortexwm ROI
Features with less than 25% nnzero in the stroke ROI of any run.

```{r}
normalized.stroke <- normalized |>
    subsetPixels(stroke == T) |>
    summarizeFeatures(stat = c("sd", "nnzero", "mean"), groups = normalized$run[normalized$stroke == T], na.rm = T, chunkopts = list(nchunks = 1)) 

normalized.cortexwm <- normalized |>
    subsetPixels(cortexwm == T) |>
    summarizeFeatures(stat = c("sd", "nnzero", "mean"), groups = normalized$run[normalized$cortexwm == T], na.rm = T, chunkopts = list(nchunks = 1)) 
```

Compute features for stroke ROI.
```{r}
normalized.stroke.features <- normalized.stroke |>
    fData() |>
    as_tibble() |>
    mutate(fid = row_number()) |>
    dplyr::rename(mean_all = mean, sd_all = sd, nnzero_all = nnzero) |>
    pivot_longer(
        cols = matches("^mouse\\."),
        names_to = c("run", ".value"),
        names_pattern = "(mouse\\.\\d+\\.\\d+)\\.(.*)"
    ) |>
        mutate(run = stringr::str_replace_all(run, "\\.", "-"))

stroke.anyrun.nnzero.lt25 <- normalized.stroke.features |> 
    left_join(normalized.stroke |>
        pData() |>
        as_tibble() |>
        group_by(run) |>
        count()
    ) |>
    mutate(prop.nnzero = nnzero/n) |>
    group_by(fid) |>
    filter(any(prop.nnzero < .25)) |>
    pull(fid) |>
    unique()
```

Compute features for cortexwm ROI
```{r}
normalized.cortexwm.features <- normalized.cortexwm |>
    fData() |>
    as_tibble() |>
    mutate(fid = row_number()) |>
    dplyr::rename(mean_all = mean, sd_all = sd, nnzero_all = nnzero) |>
    pivot_longer(
        cols = matches("^mouse\\."),
        names_to = c("run", ".value"),
        names_pattern = "(mouse\\.\\d+\\.\\d+)\\.(.*)"
    ) |>
        mutate(run = stringr::str_replace_all(run, "\\.", "-"))

cortexwm.anyrun.nnzero.lt25 <- normalized.cortexwm.features |> 
    left_join(normalized.cortexwm |>
        pData() |>
        as_tibble() |>
        group_by(run) |>
        count()
    ) |>
    mutate(prop.nnzero = nnzero/n) |>
    group_by(fid) |>
    filter(any(prop.nnzero < .25)) |>
    pull(fid) |>
    unique()
```

Subset out these features.
```{r}
ind.features.drop <- sort(unique(c(stroke.anyrun.nnzero.lt25, cortexwm.anyrun.nnzero.lt25, mean.sd.p1)))

normalized <- normalized |>
                subsetFeatures(!(1:length(mz(normalized)) %in% ind.features.drop))

```

```{r}
writeMSIDataSafe(normalized, file.path(preprocessing.data.path, "IS-normalized-filtered.imzML"))
```

# Step 3: Statistical modeling

## Subset ROI of interest
```{r}
modeling.mse <- readMSIData(file.path(preprocessing.data.path, "IS-normalized-filtered.imzML")) |>
    subsetPixels(stroke | cortexwm)
```


## Assign experimental variables
```{r}
modeling.mse$day <- factor(sub("mouse-(\\d+)-\\d+", "\\1", modeling.mse$run), levels = c(7, 28))

# we need to rename the rep so that the model know mouse 1 from day=7 is a different mouse from mouse 1 day 28

reps <- as.integer(sub("mouse-\\d+-(\\d+)", "\\1", modeling.mse$run))

modeling.mse$subject<- as.factor(ifelse(modeling.mse$day == 28, reps, reps + 5))

# lets also make a new factor for tissue

# confirm no overlaps with cortexwm and stroke, should be False
cat("Any pixels labeled as both cortex white matter AND stroke: ", 
    any(modeling.mse$stroke & modeling.mse$cortexwm), 
    "\n")

modeling.mse$tissue <- factor(ifelse(modeling.mse$stroke, "stroke", "cortexwm"),
                            levels = c("cortexwm", "stroke", "other"))

modeling.mse$samples <- factor(interaction(modeling.mse$subject, modeling.mse$tissue))
```


```{r}
modeling.mse |>
    pData() |>
    as_tibble() |>
    group_by(day, subject, tissue) |>
    summarise(nsamples = length(unique(samples)))

modeling.mse |>
    pData() |>
    as_tibble() |>
    group_by(day, subject, tissue, samples) |>
    summarise(pixels.to.be.avgd = n())

# all nsamples should be 1   
```

## Specify statisical model and fit with meansTest
```{r}
mt <- meansTest(modeling.mse, ~ day * tissue, ~ 1|subject, samples = modeling.mse$samples, use_lmer = T, na.rm = T)
```

## Visualize diagnostic plots

### QQ plots
```{r}
qq_plot_frame <- NULL
set.seed(2025)
for (fid in 1:length(mt)) {
  model <- mt[[fid]]
  mf <- model.frame(model)
  sample <- rownames(model.frame(model))
  rownames(mf) <- NULL
  qq_plot_frame <- bind_rows(qq_plot_frame,
                        cbind(tibble(feature_id = fid,
                               residuals = residuals(model),
                               fitted = fitted(model),
                               sample = sample),
                               mf))
} 
```

```{r}
qq_plot_frame <-
  qq_plot_frame |>
  #mutate(feature_id = as.numeric(feature_id)) |>
  left_join(
    tibble(mz = mz(modeling.mse)) |>
      mutate(feature_id = row_number(),
             mz = round(mz, 4)),
    by = join_by("feature_id")
  ) |>
  group_by(mz) |>
  arrange(residuals) |>
  mutate(
    samp = residuals,
    theo = qnorm(ppoints(n()))
  ) |>
  ungroup()

fids <- unique(qq_plot_frame$feature_id)
```

```{r}
manage_directory("QQ-Itx-SII-Plots-Stroke-TissuexDays")

mse.norm.filt <- readMSIData(file.path(preprocessing.data.path, "IS-normalized-filtered.imzML"))
mse.norm.filt$day <- factor(sub("mouse-(\\d+)-\\d+", "\\1", mse.norm.filt$run), levels = c(7, 28))

# we need to rename the rep so that the model know mouse 1 from day=7 is a different mouse from mouse 1 day 28

reps <- as.integer(sub("mouse-\\d+-(\\d+)", "\\1", mse.norm.filt$run))

mse.norm.filt$subject<- as.factor(ifelse(mse.norm.filt$day == 28, reps, reps + 5))

mse.norm.filt$tissue <- factor(ifelse(mse.norm.filt$stroke, "stroke", ifelse(mse.norm.filt$cortexwm, "cortexwm", "other")),
                            levels = c("cortexwm", "stroke", "other"))

mse.norm.filt$samples <- factor(interaction(mse.norm.filt$subject, mse.norm.filt$tissue))
# keep in mind that the models are run on just just stroke and cortexwm tissue so 1.other, etc. doesn't exist


# ---- SII prep (built once); carry full tissue factor for 3-zone alpha ----
pdata_sii <- as.data.table(as.data.frame(pData(mse.norm.filt)))[
  , .(x, y, run, tissue, day)]

run_levels <- c(paste0("mouse-7-", 1:5), paste0("mouse-28-", 1:5))

mz_lookup <- qq_plot_frame |>
  dplyr::distinct(feature_id, mz) |>
  dplyr::arrange(feature_id)

annot_lookup <- lcms.peaks |>
  dplyr::transmute(mz = round(`Input Mass`, 4), lipid = `LIPID MAPS`) |>
  dplyr::distinct(mz, .keep_all = TRUE)

# ---- Panel C: single-ion images; stroke + cortexwm full brightness, other dimmed ----
sii <- function(fid, roi_alpha = 1, other_alpha = 0.3, gray_other = FALSE) {
  target_mz <- mz_lookup$mz[mz_lookup$feature_id == fid]
  fi   <- which.min(abs(mz(mse.norm.filt) - target_mz))   # exact at 4 dp
  vals <- as.numeric(intensity(mse.norm.filt)[fi, ])      # one intensity per pixel

  df <- copy(pdata_sii)
  df[, intensity := vals]
  df[, run := factor(run, levels = run_levels)]

  common <- list(
    coord_equal(),
    theme_void(),
    facet_wrap(~ run, ncol = 5),                          # row1 = day 7, row2 = day 28
    scale_fill_viridis_c(option = "turbo", na.value = "#bdbdbd"),
    theme(legend.position = "bottom",
          panel.spacing.x = unit(0, "pt"),
          panel.spacing.y = unit(1, "pt")),
    labs(fill = "Intensity")
  )

  if (gray_other) {
    df[tissue == "other", intensity := NA_real_]          # non-ROI -> gray
    ggplot(df, aes(x = x, y = -y, fill = intensity)) + geom_raster() + common
  } else {
    df[, alpha_val := fifelse(tissue %in% c("stroke", "cortexwm"),
                              roi_alpha, other_alpha)]
    ggplot(df, aes(x = x, y = -y, fill = intensity)) +
      geom_tile(aes(alpha = alpha_val)) +
      scale_alpha_identity(guide = "none") + common
  }
}

# ---- Panel B: paired intensity across tissue, per-day mean overlay, subject labels ----
itx_fancy <- function(tmp_df) {
  df <- tmp_df |>
    dplyr::mutate(tissue = factor(tissue, levels = c("cortexwm", "stroke")),
                  day    = factor(day, levels = c("7", "28")))

  ggplot(df, aes(x = tissue, y = intensity, group = subject,
                 color = day, fill = day)) +
    ylab("Mean Intensity") +
    xlab(NULL) +
    theme_classic() +
    theme(legend.position = "bottom") +
    stat_summary(aes(group = day, color = day), fun = mean, geom = "point",
                 shape = 21, size = 3.5, fill = "white", stroke = 1.5, alpha = 0.6) +
    stat_summary(aes(group = day, color = day), fun = mean, geom = "line",
                 linewidth = 3.25, alpha = 0.3, lineend = "round") +
    geom_point() +
    geom_line() +
    geom_text_repel(data = df |> dplyr::filter(tissue == "stroke"),
                    aes(label = subject), color = "black",
                    hjust = -0.4, vjust = 0.5, size = 2.5, segment.alpha = 0) +
    labs(color = "Day", fill = "Day")
}
```



```{r}
library(ggpubr)
library(ggrepel)

for (f in fids) {

  tmp_df <- qq_plot_frame |> filter(feature_id == f)

  # A: QQ of residuals — color = day, shape = tissue
  qqs <- ggplot(tmp_df, aes(theo, samp, shape = tissue, color = day)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0) +
    labs(x = "Theoretical (Normal) Quantiles", y = "Residual Quantiles",
         shape = "Tissue", color = "Day") +
    theme_linedraw() +
    theme(legend.position = "bottom", legend.box = "horizontal") +
    guides(
      color = guide_legend(direction = "vertical", ncol = 1, byrow = TRUE,
                           title.position = "left"),
      shape = guide_legend(direction = "vertical", ncol = 1, byrow = TRUE,
                           title.position = "left")
    ) +
    theme(legend.key.width = unit(0.4, "cm"), legend.spacing.x = unit(0.1, "cm"))

  # D: residuals by each IV (day, tissue)
  var_day <- ggplot(tmp_df, aes(y = day, x = residuals)) +
    geom_boxplot(aes(group = day), outlier.shape = NA, color = "black") +
    geom_point(aes(shape = tissue, color = day), size = 3, alpha = 0.7) +
    labs(x = NULL, y = "Day", shape = "Tissue", color = "Day") +
    theme_linedraw() +
    theme(legend.position = "none",
          axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

  var_tissue <- ggplot(tmp_df, aes(y = tissue, x = residuals)) +
    geom_boxplot(aes(group = tissue), outlier.shape = NA, color = "black") +
    geom_point(aes(shape = tissue, color = day), size = 3, alpha = 0.7) +
    labs(x = NULL, y = "Tissue", shape = "Tissue", color = "Day") +
    theme_linedraw() +
    theme(legend.position = "none",
          axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

  var_boxplots <- ggarrange(var_day, var_tissue, ncol = 2) |>
    annotate_figure(top = "Distribution of residuals by independent variable",
                    bottom = "Residuals",
                    fig.lab = "D", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  qqs <- annotate_figure(
      qqs, top = "Quantile-quantile plots of residuals",
      fig.lab = "A", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  boxps <- annotate_figure(
      itx_fancy(tmp_df), top = "Mean intensity by tissue and day",
      fig.lab = "B", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  sii_plot <- annotate_figure(
      sii(f), top = "Single ion images with stroke ROI highlighted",
      fig.lab = "C", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  # bottom caption: m/z + LIPID MAPS annotation
  this_mz <- round(unique(tmp_df$mz)[1], 4)
  lipid   <- annot_lookup$lipid[match(this_mz, annot_lookup$mz)]
  cap <- if (is.na(lipid) || lipid == "") {
    paste0("Feature: ", round(this_mz, 3), " m/z")
  } else {
    paste0("Feature: ", round(this_mz, 3), " m/z   |   ", lipid)
  }

  ggarr <- ggarrange(
      ggarrange(qqs, boxps, nrow = 2),
      ggarrange(sii_plot, var_boxplots, nrow = 2, heights = c(5, 3)),
      ncol = 2, widths = c(1, 2)) |>
    annotate_figure(bottom = cap)

  ggsave(file.path("QQ-Itx-SII-Plots-Stroke-TissuexDays", paste0("QQ-", f, ".png")),
         plot = ggarr, height = 9, width = 11, units = "in")
}
```


# Step 4: Statistical inference

```{r}

ct <- contrastTest(mt, specs= ~ day * tissue, 
                      method="pairwise")

topFeatures(ct) |> 
    as_tibble() |>
    ggplot(aes(x = `day28...day7.estimate`, y = -log(`day28...day7.fdr`))) +
    geom_point(aes(color = `day28...day7.fdr` < 0.05), alpha = 0.5)

vol <- NULL
for (i in 1:length(mt)) {
    td <- tidy(mt[[i]])
    vol <- rbind(vol,
        tibble(i = i, est.ref = td$estimate[1], est.eff = td$estimate[2], pct.delta = (td$estimate[1] - sum(td$estimate))/sum(td$estimate), p.value = td$p.value[[2]])
    )
}

library(ggrepel)

vol |> 
mutate(fid = row_number(),
        mz = round(mz(modeling.mse), 4)) |>
left_join(lcms.peaks |> mutate(mz = round(`Input Mass`, 4))) |>
mutate(fdr = p.adjust(p.value, method = "fdr")) |>
ggplot(aes(x = pct.delta*100, y = -log(fdr))) +
    geom_point(aes(color = fdr < 0.01), alpha = 0.5) +
    geom_text_repel(
        data = ~ slice_min(.x, fdr, n = 25, with_ties = FALSE),
        aes(label = `LIPID MAPS`),
        size = 3, max.overlaps = Inf
    ) +
    labs(x = "In stroke: 7 Days - 28 Days, percent difference from 7 days")

```


```{r}
# ==============================================================================
# Statistical inference (volcano plots) and future-experiment design
# (power curves) for a day * tissue mixed-effects MSI experiment.
#
# Hypothesis testing uses Cardinal::contrastTest (vignette 4, "Method 1").
#
# Verified design (mt = MeansTest, 125 features):
#   intensity ~ day + tissue + day:tissue + (1 | subject)
#     day    : "7" (ref), "28"            -> BETWEEN-subject (5 subjects/day group)
#     tissue : "cortexwm" (ref), "stroke" -> WITHIN-subject  (both tissues/subject)
#
# Hypotheses (both are pairwise contrasts already present in contrastTest output):
#   H1  In stroke:  day7 - day28      == "day7 stroke - day28 stroke"      (between)
#   H2  In day28:   cortexwm - stroke == "day28 cortexwm - day28 stroke"   (within)
#
# Planning variance inputs (medians across the 125 features):
#   median sigma2_subject ~ 851.6,  median sigma2_residual ~ 3566.9
#   (mt[[1]] is a high-variance outlier at 9203 / 20814.)
#
# Style mirrors:
#   .../4-Statistical-Inference.html  and  .../5-Designing-Future-Experiments.html
# ==============================================================================

library(tidyverse)
library(Cardinal)
library(lmerTest)   # provides fixef()/VarCorr() for the reference means & variances

# Hypothesis labels (facet titles / colour keys / join keys).
lab_h1 <- "In stroke: day 7 - day 28"      # between-subject
lab_h2 <- "In day 28: cortexwm - stroke"   # within-subject

# Exact contrast names as emitted by contrastTest (see your ct[[1]] output).
name_h1 <- "day7 stroke - day28 stroke"
name_h2 <- "day28 cortexwm - day28 stroke"


# ==============================================================================
# PART 1 -- STATISTICAL INFERENCE via Cardinal::contrastTest
# ==============================================================================

# Run all pairwise contrasts (or reuse the `ct` you already computed).
ct <- contrastTest(mt, specs = ~ day * tissue, method = "pairwise")

# Wide results table; topFeatures applies FDR to each contrast independently.
tf <- as_tibble(topFeatures(ct))   # returns all features; pass n = to limit

# topFeatures column names are make.names(contrast) + "." + {estimate,pvalue,fdr}.
# Pull one contrast's columns and strip the prefix back to estimate/pvalue/fdr.
pull_contrast <- function(tf, contrast_name, label) {
  prefix <- paste0(make.names(contrast_name), ".")
  if (!any(startsWith(names(tf), prefix)))
    stop("No columns start with '", prefix, "'.\nAvailable: ",
         paste(names(tf), collapse = ", "))
  tf |>
    select(i, mz, starts_with(prefix)) |>
    rename_with(~ str_remove(.x, fixed(prefix)), starts_with(prefix)) |>
    mutate(contrast = label)
}

# Reference-level cell means from the fixed effects (reference levels = day7,
# cortexwm), used only to express the estimate as a percent difference.
#   day-7 stroke   = (Intercept) + tissuestroke
#   day-28 cortexwm = (Intercept) + day28
ref_means <- bind_rows(lapply(seq_len(length(mt)), function(k) {
  b <- fixef(mt[[k]])
  tibble(i      = k,
         ref_h1 = unname(b["(Intercept)"] + b["tissuestroke"]),
         ref_h2 = unname(b["(Intercept)"] + b["day28"]))
}))

signift_stroke <- bind_rows(
  pull_contrast(tf, name_h1, lab_h1) |>
    left_join(ref_means |> transmute(i, ref = ref_h1), by = "i"),
  pull_contrast(tf, name_h2, lab_h2) |>
    left_join(ref_means |> transmute(i, ref = ref_h2), by = "i")
) |>
  mutate(pct_delta = estimate / ref,
         signif_after_fdr_adj = fdr < 0.1) |>   # threshold adjustable
  arrange(pvalue)

# Significant after FDR adjustment, per hypothesis.
signift_stroke |>
  summarize(Count = n(), .by = c(contrast, signif_after_fdr_adj)) |>
  knitr::kable()

signift_stroke |>
  ggplot(aes(x = pct_delta, y = -log10(pvalue), color = signif_after_fdr_adj)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "gray40") +
  #geom_vline(xintercept = 0, color = "gray80") +
  facet_wrap(~contrast) +
  scale_x_continuous(labels = scales::percent) +
  scale_color_manual(
    values = c("TRUE" = "#E41A1C", "FALSE" = "gray50"),
    labels = c("TRUE" = "Significant at FDR < 0.1", "FALSE" = "Not significant"),
    name   = "FDR-adjusted"
  ) +
  labs(
    x = "Estimated Percent Difference (relative to reference level)",
    y = expression(-log[10](italic(P)~value))
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom"
  )

ggsave(file.path("Recreated-Figures", "Figure-Stroke-Volcano.png"),
       height = 5.75, width = 6, units = "in")

# ---- (Optional) unadjusted p-value distributions (SI, as in vignette 4) ------
signift_stroke |>
  ggplot(aes(x = pvalue)) +
  geom_histogram(bins = 7, fill = "gray50", color = "white", boundary = 0) +
  facet_wrap(~contrast) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.05)) +
  labs(x = expression(italic(P)~value), y = "Count") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white", color = "black"),
        strip.text = element_text(face = "bold"))


# ==============================================================================
# PART 2 -- DESIGNING FUTURE EXPERIMENTS: power / sample-size curves
# (adapts vignette 5; variance-component extraction, not hypothesis testing)
# ==============================================================================

df_var <- bind_rows(lapply(seq_len(length(mt)), function(i) {
  m  <- mt[[i]]
  vc <- as.data.frame(VarCorr(m))
  tibble(sigma2_subject  = vc$vcov[1],                 # subject intercept var
         sigma2_residual = attr(VarCorr(m), "sc")^2)   # residual var
}))

FDR   <- 0.05
power <- 0.9

median.sigma.subject <- median(df_var$sigma2_subject,  na.rm = TRUE)  # ~851.6
median.sigma.error   <- median(df_var$sigma2_residual, na.rm = TRUE)  # ~3566.9

# One percent-difference denominator per hypothesis (baselines differ ~2.3x).
ref_by_comparison <- signift_stroke |>
  group_by(contrast) |>
  summarize(ref = median(ref, na.rm = TRUE), .groups = "drop")

# Normal-approximation sample size (verbatim from vignette 5).
calc_sample_size <- function(desired_difference, alpha, power,
                             variance_estimate_without_I) {
  delta   <- seq(desired_difference[1], desired_difference[2], by = 1)
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta  <- qnorm(power)
  aa      <- z_alpha + z_beta
  numSample <- variance_estimate_without_I / (delta / aa)^2
  data.frame(delta, numSample, FDR, power)
}

alpha_1  <- power * FDR / (1 + (1 - FDR) * 1)    # single comparison
alpha_50 <- power * FDR / (1 + (1 - FDR) * 50)   # FDR, 50% true-negative rate
alpha_90 <- power * FDR / (1 + (1 - FDR) * 90)   # FDR, 90% true-negative rate

# Variance of the estimated difference:
#   H2 within-subject : 2 * residual
#   H1 between-subject: 2 * (residual + subject)
var_within  <- 2 * (median.sigma.error)
var_between <- 2 * (median.sigma.error + median.sigma.subject)

comparison_labels <- c(lab_h2, lab_h1)   # within first (lighter), between second
delta_range <- c(1, 1500)

df_curves <- bind_rows(
  calc_sample_size(delta_range, alpha_90, power, var_between) |>
    mutate(contrast = lab_h1, n_sig_test = 90),
  calc_sample_size(delta_range, alpha_90, power, var_within)  |>
    mutate(contrast = lab_h2, n_sig_test = 90),
  calc_sample_size(delta_range, alpha_50, power, var_between) |>
    mutate(contrast = lab_h1, n_sig_test = 50),
  calc_sample_size(delta_range, alpha_50, power, var_within)  |>
    mutate(contrast = lab_h2, n_sig_test = 50),
  calc_sample_size(delta_range, alpha_1,  power, var_between) |>
    mutate(contrast = lab_h1, n_sig_test = 1),
  calc_sample_size(delta_range, alpha_1,  power, var_within)  |>
    mutate(contrast = lab_h2, n_sig_test = 1)
) |>
  left_join(ref_by_comparison, by = "contrast") |>
  mutate(
    pct        = delta / ref,
    comparison = factor(contrast, levels = comparison_labels),
    n_sig_test = factor(n_sig_test, levels = c(1, 50, 90),
                        labels = c("No correction, i.e. a single comparison",
                                   "FDR correction, 50% true negative rate",
                                   "FDR correction, 90% true negative rate"))
  )

# ---- Power curves (styled as in vignette 5) ---------------------------------
greys <- setNames(c("grey70", "grey30"), c(lab_h2, lab_h1))  # within light / between dark

ggplot(df_curves, aes(x = pct, y = numSample, color = comparison)) +
  geom_line(aes(linetype = n_sig_test), linewidth = 0.9) +
  geom_hline(yintercept = 5, alpha = 0.5, linewidth = 0.7) +   # study replication
  scale_linetype_manual(values = c("solid", "31", "11"),
                        name = "Multiple Testing Correction") +
  scale_color_manual(values = greys) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0.02, 0.35)) +
  ylim(0, 20) +
  labs(x = "Absolute Value of Anticipated Percent Difference of Mean Intensities",
       y = "Minimum Number of Biological Replicates",
       color = "Comparison") +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(color    = guide_legend(order = 1, nrow = 2,
                                 theme = theme(legend.title.position = "top")),
         linetype = guide_legend(order = 2, nrow = 3,
                                 theme = theme(legend.title.position = "top")))

ggsave(file.path("Recreated-Figures", "Figure-Stroke-Power.png"),
       height = 6, width = 6, units = "in")

# ---- Effect size that keeps each comparison near the study's replication (5) --
df_curves |>
  filter(n_sig_test == "FDR correction, 90% true negative rate") |>
  group_by(comparison) |>
  filter(abs(numSample - 5) == min(abs(numSample - 5))) |>
  mutate(Percent_Change = pct * 100) |>
  knitr::kable()
```




# Step 5: Designing future experiments

# ANALYSIS: between stroke day conditions

# Step 2: Filtering and aggregation
## Non-specific Filtering

```{r}
normalized <- readMSIData(file.path(preprocessing.data.path, "IS-normalized.imzML"))

intensity(normalized) = as.matrix(intensity(normalized))
```

Filter marker features and IS
```{r}
normalized <- normalized |>
            subsetFeatures(!isMarker & !isIS)
```

### Identify low mean and low sd features
Features in the 10th percentile or less for both mean and sd.

```{r}
normalized <- normalized |>
    summarizeFeatures(stat = c("sd", "nnzero", "mean"), na.rm = T, chunkopts = list(nchunks = 1))

mean.sd.p1 <- normalized |>
    fData() |>
    as_tibble() |>
    mutate(fid = row_number()) |>
    filter(mean < quantile(mean, p = .1) & sd < quantile(sd, p = .1)) |>
    pull(fid) |>
    unique()
```

### Identify features with sparse stroke ROI
Features with less than 24% nnzero in the stroke ROI of any run.

```{r}
normalized.stroke <- normalized |>
    subsetPixels(stroke == T) |>
    summarizeFeatures(stat = c("sd", "nnzero", "mean"), groups = normalized$run[normalized$stroke == T], na.rm = T, chunkopts = list(nchunks = 1)) 

k_long <- normalized.stroke |>
    fData() |>
    as_tibble() |>
    mutate(fid = row_number()) |>
    dplyr::rename(mean_all = mean, sd_all = sd, nnzero_all = nnzero) |>
    pivot_longer(
        cols = matches("^mouse\\."),
        names_to = c("run", ".value"),
        names_pattern = "(mouse\\.\\d+\\.\\d+)\\.(.*)"
    ) |>
        mutate(run = stringr::str_replace_all(run, "\\.", "-"))

anyrun.nnzero.lt25 <- k_long |> 
    left_join(normalized.stroke |>
        pData() |>
        as_tibble() |>
        group_by(run) |>
        count()
    ) |>
    mutate(prop.nnzero = nnzero/n) |>
    group_by(fid) |>
    filter(any(prop.nnzero < .25)) |>
    pull(fid) |>
    unique()

```

Subset out these features.
```{r}
ind.features.drop <- sort(unique(c(anyrun.nnzero.lt25, mean.sd.p1)))

normalized <- normalized |>
                subsetFeatures(!(1:length(mz(normalized)) %in% ind.features.drop))

```


# Step 3: Statistical modeling

## Assign experimental variables
```{r}
normalized$day <- factor(sub("mouse-(\\d+)-\\d+", "\\1", normalized$run), levels = c(7, 28))

normalized$rep<- as.factor(sub("mouse-\\d+-(\\d+)", "\\1", normalized$run))
```

## Subset ROI of interest
```{r}
modeling.mse <- normalized |>
    subsetPixels(stroke == TRUE)
```

## Specify statisical model and fit with meansTest
```{r}
mt <- meansTest(modeling.mse, ~ day, samples = modeling.mse$run)
```

## Visualize diagnostic plots

### QQ plots
```{r}
qq_plot_frame <- NULL
set.seed(2025)
for (fid in 1:length(mt)) {
  model <- mt[[fid]]
  mf <- model.frame(model)
  run <- rownames(model.frame(model))
  rownames(mf) <- NULL
  qq_plot_frame <- bind_rows(qq_plot_frame,
                        cbind(tibble(feature_id = fid,
                               residuals = residuals(model),
                               fitted = fitted(model),
                               run = run),
                               mf))
} 
```

```{r}
qq_plot_frame <-
  qq_plot_frame |>
  #mutate(feature_id = as.numeric(feature_id)) |>
  left_join(
    tibble(mz = mz(modeling.mse)) |>
      mutate(feature_id = row_number(),
             mz = round(mz, 4)),
    by = join_by("feature_id")
  ) |>
  group_by(mz) |>
  arrange(residuals) |>
  mutate(
    samp = residuals,
    theo = qnorm(ppoints(n()))
  ) |>
  ungroup()

fids <- unique(qq_plot_frame$feature_id)
```

```{r}
manage_directory("QQ-Itx-SII-Plots-Stroke")

# ---- SII prep (built once) ----
pdata_sii <- as.data.table(as.data.frame(pData(normalized)))[
  , .(x, y, run, stroke, hemisphere, day, rep)]

run_levels <- c(paste0("mouse-7-", 1:5), paste0("mouse-28-", 1:5))

mz_lookup <- qq_plot_frame |>
  dplyr::distinct(feature_id, mz) |>
  dplyr::arrange(feature_id)

# ---- Annotation lookup: mz (4 dp) -> LIPID MAPS ----
annot_lookup <- lcms.peaks |>
  dplyr::transmute(mz = round(`Input Mass`, 4),
                   lipid = `LIPID MAPS`) |>
  dplyr::distinct(mz, .keep_all = TRUE)   # first hit if an Input Mass repeats

# ---- Panel C: single-ion images, stroke hemisphere only, vertically stacked ----
sii <- function(fid, stroke_alpha = 1, nonstroke_alpha = 0.3, gray_nonstroke = FALSE) {
  target_mz <- mz_lookup$mz[mz_lookup$feature_id == fid]
  fi   <- which.min(abs(mz(normalized) - target_mz))     # nearest row (exact at 4 dp)
  vals <- as.numeric(intensity(normalized)[fi, ])        # one intensity per pixel

  df <- copy(pdata_sii)
  df[, intensity := vals]
  df <- df[hemisphere == TRUE]                           # stroke hemisphere only
  df[, run := factor(run, levels = run_levels)]

  common <- list(
    coord_equal(),
    theme_void(),
    facet_wrap(~ run, ncol = 2, dir = "v"),              # col1 = day 7, col2 = day 28
    scale_fill_viridis_c(option = "turbo", na.value = "#bdbdbd"),
    theme(legend.position = "bottom",
          panel.spacing.x = unit(0, "pt"),
          panel.spacing.y = unit(1, "pt")),
    labs(fill = "Intensity")
  )

  if (gray_nonstroke) {
    df[stroke == FALSE, intensity := NA_real_]           # non-stroke -> gray
    ggplot(df, aes(x = x, y = -y, fill = intensity)) + geom_raster() + common
  } else {
    df[, alpha_val := fifelse(stroke, stroke_alpha, nonstroke_alpha)]
    ggplot(df, aes(x = x, y = -y, fill = intensity)) +
      geom_tile(aes(alpha = alpha_val)) +
      scale_alpha_identity(guide = "none") + common
  }
}

# ---- Panel B: intensity by day, larger points + mean marker, no lines ----
itx_fancy <- function(tmp_df) {
  jit <- position_jitter(width = 0.08, height = 0, seed = 1)
  ggplot(tmp_df, aes(x = day, y = intensity, color = day)) +
    ylab("Intensity") +
    xlab("Day") +
    theme_classic() +
    theme(legend.position = "bottom") +
    geom_point(position = jit, size = 2.5, alpha = 0.8) +
    geom_text_repel(aes(label = run), position = jit,
                    color = "black", size = 2.5,
                    segment.alpha = 0.3, max.overlaps = Inf,
                    show.legend = FALSE) +
    stat_summary(aes(group = day), fun = mean, geom = "point",
                 shape = 21, size = 6, fill = "white", stroke = 1.5, alpha = 0.6) +
    labs(color = "Day")
}
```



```{r}
library(ggpubr)
library(ggrepel)

for (f in fids[1:5]) {

  tmp_df <- qq_plot_frame |> filter(feature_id == f)

  # A: QQ of residuals (labeled)
  qqs <- ggplot(tmp_df, aes(theo, samp, color = day)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_text_repel(aes(label = run), color = "black", size = 2.5,
                    segment.alpha = 0.3, max.overlaps = Inf, show.legend = FALSE) +
    geom_abline(slope = 1, intercept = 0) +
    labs(x = "Theoretical (Normal) Quantiles", y = "Residual Quantiles", color = "Day") +
    theme_linedraw() +
    theme(legend.position = "bottom")

  # D: residual distribution by day (labeled)
  var_day <- ggplot(tmp_df, aes(y = day, x = residuals)) +
    geom_boxplot(aes(group = day), outlier.shape = NA, color = "black") +
    geom_point(aes(color = day), size = 3, alpha = 0.7) +
    geom_text_repel(aes(label = run), color = "black", size = 2.5,
                    segment.alpha = 0.3, max.overlaps = Inf, show.legend = FALSE) +
    labs(x = NULL, y = "Day", color = "Day") +
    theme_linedraw() +
    theme(legend.position = "none",
          axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

  var_boxplots <- annotate_figure(
      var_day, top = "Distribution of residuals by day", bottom = "Residuals",
      fig.lab = "D", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  qqs <- annotate_figure(
      qqs, top = "Quantile-quantile plots of residuals",
      fig.lab = "A", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  boxps <- annotate_figure(
      itx_fancy(tmp_df), top = "Intensity by day",
      fig.lab = "B", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  sii_plot <- annotate_figure(
      sii(f), top = "Single ion images with stroke ROI highlighted",
      fig.lab = "C", fig.lab.size = 20, fig.lab.face = "bold") +
    theme(panel.border = element_rect(colour = "#999999", fill = NA, linewidth = 0.25))

  # bottom caption: m/z + LIPID MAPS annotation
  this_mz <- round(unique(tmp_df$mz)[1], 4)
  lipid   <- annot_lookup$lipid[match(this_mz, annot_lookup$mz)]
  cap <- if (is.na(lipid) || lipid == "") {
    paste0("Feature: ", round(this_mz, 3), " m/z")
  } else {
    paste0("Feature: ", round(this_mz, 3), " m/z   |   ", lipid)
  }

  # stack the three non-SII charts: B (2) over A (2) over D (1); C beside
  stacked <- ggarrange(boxps, qqs, var_boxplots, nrow = 3, heights = c(2, 2, 1))
  ggarr <- ggarrange(stacked, sii_plot, ncol = 2, widths = c(1, 1.2)) |>
    annotate_figure(bottom = cap)

  ggsave(file.path("QQ-Itx-SII-Plots-Stroke", paste0("QQ-", f, ".png")),
         plot = ggarr, height = 9, width = 11, units = "in")
}
```



```{r}

ct <- contrastTest(mt, specs="day", 
                      method="pairwise")

topFeatures(ct) |> 
    as_tibble() |>
    ggplot(aes(x = `day28...day7.estimate`, y = -log(`day28...day7.fdr`))) +
    geom_point(aes(color = `day28...day7.fdr` < 0.05), alpha = 0.5)

vol <- NULL
for (i in 1:length(mt)) {
    td <- tidy(mt[[i]])
    vol <- rbind(vol,
        tibble(i = i, est.ref = td$estimate[1], est.eff = td$estimate[2], pct.delta = (td$estimate[1] - sum(td$estimate))/sum(td$estimate), p.value = td$p.value[[2]])
    )
}

library(ggrepel)

vol |> 
mutate(fid = row_number(),
        mz = round(mz(modeling.mse), 4)) |>
left_join(lcms.peaks |> mutate(mz = round(`Input Mass`, 4))) |>
mutate(fdr = p.adjust(p.value, method = "fdr")) |>
ggplot(aes(x = pct.delta*100, y = -log(fdr))) +
    geom_point(aes(color = fdr < 0.01), alpha = 0.5) +
    geom_text_repel(
        data = ~ slice_min(.x, fdr, n = 25, with_ties = FALSE),
        aes(label = `LIPID MAPS`),
        size = 3, max.overlaps = Inf
    ) +
    labs(x = "In stroke: 7 Days - 28 Days, percent difference from 7 days")

```

# Step 4: Statistical inference
# Step 5: Designing future experiments