---
title: "6-Mouse-Stroke-Case-Study"
author: "Ethan Rogers"
date: "06/29/2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

In this vignette we demonstrate the proposed workflow on a case study of mouse brain samples after photothrombotic stroke.

# Step 0: Setup

Please follow the directions in vignette `0-Setup.Rmd` to create the environment and directory trees.

# Step 1: Preprocessing

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(Cardinal)

source("Utilities.R")
source("SpatialUtilities.R")

# from wd
data.path <- file.path("Mouse-Stroke-Data")
raw.data.path <- file.path(data.path ,"Raw")
preprocessing.data.path <- file.path(data.path ,"Preprocessing")
recalib.data.path <- file.path(preprocessing.data.path,"Recalibrated")
aligned.data.path <- file.path(preprocessing.data.path,"Aligned")

outputs.path <- file.path("Outputs")
```

## Load metadata
```{r}
name_tibble <- read_csv(file.path(outputs.path, "mouse_stroke_metadata.csv"), show_col_types = FALSE)
```

## Loading data

```{r}
samp <- name_tibble |>
    filter(!is.na(days) & !is.na(replicate))
```

```{r}
msa.raw <- list()

for (s in samp$filename) {
    m <- readMSIData(file.path(raw.data.path, s))

    # reorient brains to be vertical
    crd <- coord(m)
    if (max(crd$x) - min(crd$x) < max(crd$y) - min(crd$y)) {
        coord(m)$x <- crd$y
        coord(m)$y <- crd$x
    }

    # rename run to be sample short name
    sn <- name_tibble |> 
        filter(filename == s) |>
        pull(short_name)

    runNames(m) <- sn
    msa.raw[[sn]] <- m
}

gc()
```

## Recalibration
These data do not require baseline removal or smoothing.

```{r}
bp <- SnowfastParam(workers = 10L, force.GC = T)

for (sn in samp$short_name[3:10]) {
    print(paste("Starting", sn, "\n"))
    est_ref_peaks <- msa.raw[[sn]] |>
		estimateReferencePeaks(BPPARAM = bp)
    msa.raw[[sn]] |>
	    recalibrate(ref = est_ref_peaks, tolerance = 10, units = "ppm") |>
        process(BPPARAM = bp) |>
        writeMSIDataSafe(file.path(recalib.data.path, paste0(sn, ".imzML")))
}

gc()
```

# Peak alignment from known peak list and IS C13 lipids

```{r}
lcms.peaks <- read_csv(file.path(outputs.path, 
                    "LCMSlipids_brain_wang_et_al_nature_metab_2025.csv"),
                    show_col_types = FALSE)

#for normalization
c13.is.lipids <- c(428.299845, 499.355465, 596.372775, 708.589835, 777.629855, 785.6367, 800.6476, 876.662865)

all.peaks.to.align <- sort(c(c13.is.lipids, lcms.peaks$`Input Mass`))
```

```{r}
bp <- SnowfastParam(workers = 8L, force.GC = T)

# load, align, save
for (sn in samp$short_name) {
    readMSIData(file.path(recalib.data.path, paste0(sn, ".imzML"))) |>
    peakAlign(ref = all.peaks.to.align) |>
    process(BPPARAM = bp) |>
    writeMSIDataSafe(file.path(aligned.data.path, paste0(sn, ".imzML")))
}
gc()
```

```{r}
ind.c13.is.lipids <- features(readMSIData(file.path(aligned.data.path, paste0(samp$short_name[1], ".imzML"))),
                                mz = c13.is.lipids)
```

```{r}
# keep the original 146 lipids

mse_list <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML")))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse <- Reduce(cbind, mse_list)

writeMSIDataSafe(mse, file.path(preprocessing.data.path, "Analytes-IS-Aligned.imzML"))
```

```{r}
# keep the original 146 lipids

mse_list <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML"))) |>
    subsetFeatures(!(1:length(all.peaks.to.align) %in% ind.c13.is.lipids)) |>
    summarizePixels(stat = c("nnzero", "mean", "sd"), na.rm = T, chunkopts = list(nchunks = 1))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse <- Reduce(cbind, mse_list)

writeMSIDataSafe(mse, file.path(preprocessing.data.path, "Analytes-Aligned.imzML"))
```

```{r}
# keep the 8 IS lipids

mse_list.is <- lapply(samp$short_name, function(short_name){
    m <- readMSIData(file.path(aligned.data.path, paste0(short_name,".imzML"))) |>
    subsetFeatures(1:length(all.peaks.to.align) %in% ind.c13.is.lipids) |>
    summarizePixels(stat = c("nnzero", "mean", "sd"), na.rm = T, chunkopts = list(nchunks = 1))
    m$x <- m$x - min(m$x) + 1
    m$y <- m$y - min(m$y) + 1
    m
})

mse.is <- Reduce(cbind, mse_list.is)

writeMSIDataSafe(mse.is, file.path(preprocessing.data.path, "IS-Aligned.imzML"))
```


## ROI selection

### Trim background 

```{r}
intensity(mse) <- as.matrix(intensity(mse))

mse <- mse |> add_edge_factor_mse(radius = 10,  contiguous = FALSE)

mse_subset <- mse |> 
                subsetPixels(((edge == F & nnzero > 78) | (edge == T & nnzero > 88)))

islands_drop_mask <- drop_pixel_islands_mask_mse(mse_subset, n = 20)

mse_subset_trim_islands <- mse_subset |> subsetPixels(islands_drop_mask)

key <- function(d) paste(d$run, d$x, d$y, sep = "\r")

tissue_mask <- key(pData(mse)) %in% key(pData(mse_subset_trim_islands))
```

```{r}
#for (i in 1:146) {
#    image_png(mse_subset_trim_islands, i = i, style = "dark", bg = "black", key = F, grid = F, layout = c(5, 2), filename = paste("tmp/feature_", i, ".png"),
#          width = 1500, height = 2200, pointsize = 24)
#}
```

### Merge annotations from QuPath

We used 834.5273 m/z (PS 40:6) to segment stroke from non-stroke area as it was previously [identified as a marker of white matter](https://doi.org/10.1038/s41598-018-37612-5) in a mouse stroke model.

```{r}
annot.qp <- read_csv(file.path(outputs.path, 
                    "stroke-hemisphere-annotations-qupath.csv"),
                    show_col_types = FALSE)
```

```{r}
# verify order (all should be true)
all(pData(mse_subset_trim_islands)$x == annot.qp$x)
all(pData(mse_subset_trim_islands)$y == annot.qp$y)
all(pData(mse_subset_trim_islands)$run == annot.qp$run)

pData(mse_subset_trim_islands)$stroke <- annot.qp$stroke
pData(mse_subset_trim_islands)$hemisphere <- annot.qp$hemisphere
```

```{r}
image(mse_subset_trim_islands, "stroke")
image(mse_subset_trim_islands, "hemisphere")
```

### Trim areas with very high internal standard lipid after tissue mask

```{r}
intensity(mse.is) <- as.matrix(intensity(mse.is))

mse.is.tmp <- mse.is |>
    subsetPixels(tissue_mask) |>
    summarizeFeatures(stat = c("nnzero", "sd", "mean"), na.rm = T) |>
    subsetFeatures(mean > 200) |>
    summarizePixels(stat = c("nnzero", "sd", "mean"), na.rm = T)

hist(mse.is.tmp$mean)

quantile(mse.is.tmp$mean, p = .99)

is.mask <- mse.is.tmp$mean < quantile(mse.is.tmp$mean, p = .99)
```

```{r}
image_png(mse.is |> subsetPixels(tissue_mask) , subset = is.mask, i = c(1,4,5,7,8), layout = c(5, 10), key = F, grid = F, style = "dark", bg = "black", filename = "is.lipids.ltp99.png", width = 10000, height = 6000, pointsize = 24)
```

```{r}
mse.roi.trimmed <- mse_subset_trim_islands |> 
    subsetPixels(is.mask)

image(mse.roi.trimmed, "stroke")
image(mse.roi.trimmed, "hemisphere")

writeMSIDataSafe(mse.roi.trimmed, file.path(preprocessing.data.path, "Aligned-Tissue-ROIs.imzML"))
```


```{r}
mse.is.roi.trimmed <- mse.is |>
    subsetPixels(tissue_mask) |>
    subsetPixels(is.mask)

writeMSIDataSafe(mse.is.roi.trimmed , file.path(preprocessing.data.path, "IS-Aligned-Tissue-ROIS.imzML"))
```

```{r}
#final checks
all(mse.roi.trimmed$run == mse.is.roi.trimmed$run)
all(mse.roi.trimmed$x == mse.is.roi.trimmed$x)
all(mse.roi.trimmed$y == mse.is.roi.trimmed$y)
```


## Normalization


```{r}
#keeping these high abundant IS lipids
c13.is.lipids[c(1,4,5,7,8)]

mse.all <- readMSIData(file.path(preprocessing.data.path, "Analytes-IS-Aligned.imzML"))

intensity(mse.all) <- as.matrix(intensity(mse.all))

mask <- key(pData(mse.all)) %in% key(pData(mse.roi.trimmed))

mse.all.trimmed <- mse.all |>
    subsetPixels(mask)

mse.all.trimmed$stroke <- mse.roi.trimmed$stroke
mse.all.trimmed$hemisphere <- mse.roi.trimmed$hemisphere

# label features
n.features <- nrow(fData(mse.all.trimmed))

fData(mse.all.trimmed)$isIS <- 1:n.features %in% features(mse.all.trimmed, mz=c13.is.lipids)

fData(mse.all.trimmed)$isMarker <- 1:n.features %in% features(mse.all.trimmed, mz=c(834.5273))
```

```{r}
normalizeMultipleRefs <- function(object, refs, scale = 1, ...) {
    if ( missing(refs) || is.null(refs) || length(refs) == 0L )
        stop("must provide one or more reference values via 'refs'")
    refs <- sort(unique(as.numeric(refs)))
    FUN <- function(x, t, refs, scale = 1, ...) {
        if ( any(refs < min(t)) || any(refs > max(t)) )
            stop("one or more 'refs' is outside of domain limits")
        i <- matter::bsearch(refs, t, nearest = TRUE)
        b <- mean(x[i], na.rm = TRUE)
        if ( !is.na(b) && b != 0 ) {
            y <- scale * x / b
        } else {
            y <- rep.int(0, length(x))
        }
        dim(y) <- dim(x)
        y
    }
    Cardinal::addProcessing(object, FUN,
        label = "intensity normalization",
        metadata = list(method = "multiref", refs = refs),
        refs = refs, scale = scale, ...)
}
```

```{r}
#drop features before normalization
prenormalized <- mse.all.trimmed |>
    summarizeFeatures() |>
    subsetFeatures(!isMarker & !(isIS & mean < 200)) 

normalized <- prenormalized |>
    normalizeMultipleRefs(mz(prenormalized)[fData(prenormalized)$isIS], scale = 1000) |>
    process()

```

```{r}
for (i in 1:length(mz(normalized))) {
    image_png(normalized, i = i, style = "dark", bg = "black", key = F, grid = F, layout = c(5, 2), filename = paste("tmp/normalized/feature_", i, ".png"),
          width = 1500, height = 2200, pointsize = 24)
}
```

```{r}
writeMSIDataSafe(mse_subset_trim_islands, file.path(preprocessing.data.path, "IS-normalized.imzML"))
```


# Step 2: Filtering and aggregation

# Step 3: Statistical modeling

```{r}

normalized$day <- factor(sub("mouse-(\\d+)-\\d+", "\\1", normalized$run), levels = c(7, 28))

normalized$rep<- as.factor(sub("mouse-\\d+-(\\d+)", "\\1", normalized$run))

modeling.mse <- normalized |>
    subsetPixels(stroke == TRUE)

mt <- meansTest(modeling.mse, ~ day, samples = modeling.mse$run)
ct <- contrastTest(mt, specs="day", 
                      method="pairwise")

topFeatures(ct) |> 
    as_tibble() |>
    ggplot(aes(x = `day28...day7.estimate`, y = -log(`day28...day7.fdr`))) +
    geom_point(aes(color = `day28...day7.fdr` < 0.05), alpha = 0.5)

vol <- NULL
for (i in 1:length(mt)) {
    td <- tidy(mt[[i]])
    vol <- rbind(vol,
        tibble(i = i, est.ref = td$estimate[1], est.eff = td$estimate[2], pct.delta = (td$estimate[1] - sum(td$estimate))/sum(td$estimate), p.value = td$p.value[[2]])
    )
}

library(ggrepel)

vol |> 
mutate(fid = row_number(),
        mz = round(mz(modeling.mse), 4)) |>
left_join(lcms.peaks |> mutate(mz = round(`Input Mass`, 4))) |>
mutate(fdr = p.adjust(p.value, method = "fdr")) |>
ggplot(aes(x = pct.delta*100, y = -log(fdr))) +
    geom_point(aes(color = fdr < 0.01), alpha = 0.5) +
    geom_text_repel(
        data = ~ slice_min(.x, fdr, n = 25, with_ties = FALSE),
        aes(label = `LIPID MAPS`),
        size = 3, max.overlaps = Inf
    ) +
    labs(x = "In stroke: 7 Days - 28 Days, percent difference from 7 days")

```

# Step 4: Statistical inference

# Step 5: Designing future experiments