---
title: "6-Simulation-1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

Simulation 1 was designed to evaluate the performance of multivariate clustering methods, speciclaky spatiual shrunken centroids and spatiual K means, for ROI segmentation. We will create images with spatially autocorrelated background noise and clearly defined ROIs. All samples will share a set of features that clearly and consistantly define regions of interest. Then there will be a set of nuisence features, which do not contain the ROI. Finally, there will be 500 features that have a differnetially abundennt effects in the ROIs only. We will then segment the ROIs with two methods and review the results.    

# Load dependencies

```{r load-dependencies, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(Cardinal)
library(broom)

source("Utilities.R")
source("SimulationUtilities.R")

RNGkind("L'Ecuyer-CMRG")
setCardinalVerbose(verbose = FALSE)
```

# Simulation overview

The simulation begins from an existing list of image objects saved as `sim_list_original`. These are the same base images that will be used for Simulation 2, and were simulated using Cardinal's `simulateImage` function.

The major ingredients are:

- a shared baseline intensity level
- feature-specific pixel standard deviations
- a sample table defining subject, tissue, and condition
- feature-level effect vectors defining where true signal exists
- residual variation
- an oracle ROI used as the reference target
- two segmentation strategies to compare against the oracle ROI

# Define the baseline image structure

We first define a feature baseline and a per-feature target standard deviation. The simulation uses 3000 features and a 100 x 100 pixel grid.

```{r baseline-setup}
baseline <- 80
set.seed(2025)
pixel_sd_base <- 50
pixel_sd <- rnorm(3000, mean = pixel_sd_base, sd = 7) + 5

if (any(pixel_sd < 5)) {
  warning("Some pixel_sd values are below 5", call. = FALSE, immediate. = TRUE)
}

pdx <- PositionDataFrame(expand_grid(x = 1:100, y = 1:100))
pdx$all <- TRUE
fdx <- MassDataFrame(1:3000)
fdx$all <- baseline
```

This simulation does enforce the assumption that not every pixel will have the same standard deviation, and that standard deviations of pixels are normally distributed. . At small values that is not possible, as sd cannot be negative - thus we will arbitrarily increase values to avoid this. We will also flag very small values for pixel sd.

# Define the sample structure

The simulation uses 16 samples, corresponding to 8 subjects crossed with a binary tissue indicator and a binary condition indicator.  We will make this design more complex than necessary to assist with our simulations later.

```{r sample-metadata}
samples <- data.frame(
  subject = c(1:8, 1:8),
  tissue = c(rep(0, 8), rep(1, 8)),
  condition = c(rep(0, 4), rep(1, 4), rep(0, 4), rep(1, 4))
) |>
  mutate(id = paste0(subject, "_", tissue, "_", condition))

knitr::kable(samples)
```

This layout is intentionally simple. The goal is not to reproduce every detail of the experimental design from the real data, but to create a structured setting where signal can be attached to condition, tissue, or both, and then tested after segmentation.

# Load the starting image objects

```{r load-simulation-base}
load("Simulation-Data/sim_list_original.RData")
length(sim_list_original)
```

# Define feature-level effects and variance components

The next block is where the important simulation assumptions are encoded. We specify:

- `delta_in_roi`, the magnitude of the condition or tissue effect in the ROI
- `ROI_def_delta`, the magnitude of the intensity difference between the background and ROI in ROI defining features
- `resid`, the residual variation scale

```{r effect-parameters}
delta_in_roi <- 30
resid <- sqrt(70)

ROI_def_delta <- 40
```

The simulation then builds feature-wise effect vectors. The condition effect is restricted to the final 500 features, and the ROI-defining features are concentrated in the first 100 features.

```{r effect-vectors}
set.seed(2024)
effects_normal <- tibble(
  mz = mz(sim_list_original[[1]]),
  condition = c(rep(0, 2500), rep(-1 * delta_in_roi, 250), rep(delta_in_roi, 250)),
  ROI_defining_features = c(rep(ROI_def_delta, 50), rep(-ROI_def_delta, 50), rep(0, 2900))
)
```

This is doing several things at once.

First, it makes only a subset of features truly differential. That is necessary if we want to talk meaningfully about both true *and* false positives later.

Second, it separates the features that define the ROI from the features that carry the condition effect. This is to mirror the real world situation that the biological structures (that one might want to segment ROIs of) have features that define them as well as change within them. The methods can use any features of the feature set to draw the ROIs.


# Generate residual variation

The script next generates a sample-by-feature residual variation matrix.

```{r variance-matrices}
set.seed(2026)
residual_mat <- matrix(rnorm(16 * nrow(effects_normal), sd = resid), nrow = 16)
residual_mat <- scale(residual_mat, center = TRUE, scale = FALSE)
```

This keeps the active simulation relatively easy to interpret: the feature-level fixed effects determine where signal should appear, while `residual_mat` controls how much those means fluctuate from sample to sample.

# Construct the target mean structures

The next step combines the fixed effects, ROI-defining features, residual variation, and baseline into mean intensity targets for the ROI and the background.

```{r target-mean-matrices}
effects_matrix_c2 <- sapply(1:16, function(i) {
  samples[[i, "condition"]] * effects_normal$condition +
    effects_normal$ROI_defining_features +
    residual_mat[i, ] +
    baseline
})

effects_matrix_bg <- sapply(1:16, function(i) {
  -1 * effects_normal$ROI_defining_features +
    residual_mat[i, ] +
    baseline
})
```

Features that define the ROI are pushed in opposite directions inside and outside the ROI. Separately, the condition effect is restricted to the final 500 features in the ROI.

A simple safety check follows.

```{r effect-sanity-check}
i <- 1
for (eff in c(effects_matrix_bg,
              effects_matrix_c2)) {
  if (any(eff < 0)) {
    stop(paste0("Some effects in ", i, " have negative means."))
  }
  i <- i + 1
}
```

Negative means would be inconsistent with real world MSI images.

# Build the oracle ROI into each image

We now define the actual pixels that make up the ROI and the background then apply the effects generated above.

```{r build-oracle-roi, message=FALSE, warning=FALSE}
sim_list_adj <- bplapply(seq_along(sim_list_original), FUN = function(i) {
  m <- sim_list_original[[i]]
  runNames(m) <- paste0("run_", i)

  intensity(m) <- scale_row_sd_mean_subset(
    as.matrix(intensity(m)),
    target_sd = pixel_sd,
    target_mean = effects_matrix_bg[, i]
  )

  pData(m) <- addShape(addShape(pData(m), c(50, 50), 20, name = "circle_1"),
                       c(50, 50), 40, name = "circle_2")

  pData(m)$circle_2 <- ifelse(pData(m)$circle_2 & pData(m)$circle_1,
                              FALSE,
                              pData(m)$circle_2)

  pData(m)$bg <- ifelse(pData(m)$circle_2, FALSE, TRUE)

  intensity(m) <- scale_row_sd_mean_subset(
    as.matrix(intensity(m)),
    target_sd = pixel_sd,
    columns_to_scale = pData(m)$circle_2,
    target_mean = effects_matrix_c2[, i]
  )

  summarizeFeatures(
    summarizeFeatures(
      m,
      stat = c("mean", "sd"),
      na.rm = TRUE,
      verbose = FALSE,
      groups = ifelse(pData(m)$circle_2, "oracle_donut", "other")
    ),
    stat = c("mean", "sd"),
    na.rm = TRUE,
    verbose = FALSE
  )
}, BPPARAM = SnowfastParam(8))
```

# Segment the simulated images with two methods

The script then applies two unsupervised segmentation methods.

The first is spatial shrunken centroids (SSC), using multiple regularization values. The second is spatial k-means (SKM), using seven clusters. Both are run separately on each simulated image.

```{r run-segmentation, message=FALSE, warning=FALSE}
sscl <- bplapply(sim_list_adj, FUN = function(m) {
  set.seed(123456789, kind = "L'Ecuyer-CMRG")
  spatialShrunkenCentroids(m, k = 4, s = c(6, 9, 12, 15), weights = "adaptive")
}, BPPARAM = SnowfastParam(8))

skml_7c <- bplapply(sim_list_adj, FUN = function(m) {
  set.seed(123456789, kind = "L'Ecuyer-CMRG")
  spatialKMeans(m, k = 7, weights = "adaptive")
}, BPPARAM = SnowfastParam(8))
```

These parameter choices are part of what the simulation is actually testing. The oracle ROI is known, so segmentation quality can be evaluated after the fact. SSC and SKM are not being used here as discovery tools in the usual sense; they are being benchmarked against a known target.

# Extract ROI summaries for modeling

After segmentation, helper functions from `SimulationUtilities.R` identify the ROI candidates, select representative clusters, and extract feature summaries for modeling.

```{r analyze-segmentation-results}
results_list <- analyze_spatial_clusters(
  sscl,
  sim_list_adj,
  samples,
  n_top_per_condition = 4,
  n_top_features = 20
)

results_list_skm_7c <- analyze_spatial_clusters_skm(
  skml_7c,
  sim_list_adj,
  samples,
  n_top_per_condition = 4,
  n_top_features = 20
)
```

These helper functions are doing a substantial amount of work. They bridge the gap between unsupervised segmentation output and a modeling-ready data frame. That is an important practical step, because in an actual analysis one never tests the clustering object directly. One tests summaries extracted from the ROI implied by that clustering.

# Fit the downstream models

The simulation compares three ROI definitions:

- the ROI recovered by spatial shrunken centroids
- the ROI recovered by spatial k-means
- the oracle donut ROI

For each, the code extracts feature-level ROI means and fits condition-only models.

```{r fit-models, results='hide', message=FALSE, warning=FALSE}
mse_frame_to_test <- extract_data_for_modeling_alt(
  results_list[["sim_to_test"]],
  results_list[["samples_to_test"]]
)

run_models_to_test_ssc_roi <- run_models_no_me(
  mse_frame_to_test,
  formu = "ssc_roi.mean ~ condition",
  verbose = FALSE
)

run_models_to_test_oracle_roi <- run_models_no_me(
  mse_frame_to_test,
  formu = "oracle_donut.mean ~ condition",
  verbose = FALSE
)

mse_frame_to_test_skm_7c <- extract_data_for_modeling_alt(
  results_list_skm_7c[["sim_to_test"]],
  results_list_skm_7c[["samples_to_test"]]
)

run_models_to_test_skm_roi_7c <- run_models_no_me(
  mse_frame_to_test_skm_7c,
  formu = "skm_roi.mean ~ condition",
  verbose = FALSE
)
```

The formula choice is intentionally simple. Once the ROI summaries have been constructed, the question is whether the recovered ROI yields the expected condition effect. By keeping the downstream model straightforward, the script places the burden of performance on the ROI definition itself.

# Compare which samples were selected for figures

The script next finds treatment and control examples that were identified by both SSC and SKM, then uses them for visualization.

```{r identify-charting-samples}
trt_sample_id <- intersect(
  results_list[[3]] |> filter(condition == 1) |> pull(id),
  results_list_skm_7c[[3]] |> filter(condition == 1) |> pull(id)
)[1]

control_sample_id <- intersect(
  results_list[[3]] |> filter(condition == 0) |> pull(id),
  results_list_skm_7c[[3]] |> filter(condition == 0) |> pull(id)
)[1]

```

This is a practical decision rather than a modeling decision. By choosing samples that are usable across both methods, the resulting figures remain directly comparable.

# Visualize the simulated images and ROI definitions

The original simulation script writes several Cardinal image panels to disk to show the ROI-defining features, a null feature, a truly differential feature, and the oracle and recovered segmentations. Those images are helpful context, so here we display the same comparisons inline while also saving them.

```{r contextual-simulation-images, fig.width=6, fig.height=12, message=FALSE, warning=FALSE}

save_cardinal_plot <- function(filename, expr, width = 600, height = 1200) {
  png(filename = file.path("Recreated-Figures", filename), width = width, height = height)
  on.exit(dev.off(), add = TRUE)
  print(eval.parent(substitute(expr)))
}

m_context <- cbind(
  sim_list_adj[[which(samples$id == control_sample_id)]],
  sim_list_adj[[which(samples$id == trt_sample_id)]]
)

ground_truth_roi_defining_plot <- image(
  m_context,
  i = c(1),
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Ground-Truth-ROI-Defining.png",
  image(
    m_context,
    i = c(1),
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
ground_truth_roi_defining_plot

ground_truth_no_effect_plot <- image(
  m_context,
  i = c(1000),
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Ground-Truth-No-Effect.png",
  image(
    m_context,
    i = c(1000),
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
ground_truth_no_effect_plot

ground_truth_diff_abun_plot <- image(
  m_context,
  i = c(2700),
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Ground-Truth-Diff-Abundance.png",
  image(
    m_context,
    i = c(2700),
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
ground_truth_diff_abun_plot

oracle_segmentation_plot <- image(
  m_context,
  "circle_2",
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Segmentation-Oracle.png",
  image(
    m_context,
    "circle_2",
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
oracle_segmentation_plot

m_ssc_context <- cbind(
  results_list[[2]][[which(results_list[[3]]$id == control_sample_id)]],
  results_list[[2]][[which(results_list[[3]]$id == trt_sample_id)]]
)

ssc_segmentation_plot <- image(
  m_ssc_context,
  "roi_ssc",
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Segmentation-SSC.png",
  image(
    m_ssc_context,
    "roi_ssc",
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
ssc_segmentation_plot

m_skm_context <- cbind(
  results_list_skm_7c[[2]][[which(results_list_skm_7c[[3]]$id == control_sample_id)]],
  results_list_skm_7c[[2]][[which(results_list_skm_7c[[3]]$id == trt_sample_id)]]
)

skm_segmentation_plot <- image(
  m_skm_context,
  "roi_skm",
  layout = c(2, 1),
  scale = TRUE,
  superpose = FALSE,
  key = FALSE,
  grid = FALSE,
  xaxt = "n",
  yaxt = "n",
  ylab = NA,
  xlab = NA
)
save_cardinal_plot(
  "Figure-ROI-Simulation-1-Segmentation-SKM.png",
  image(
    m_skm_context,
    "roi_skm",
    layout = c(2, 1),
    scale = TRUE,
    superpose = FALSE,
    key = FALSE,
    grid = FALSE,
    xaxt = "n",
    yaxt = "n",
    ylab = NA,
    xlab = NA
  )
)
skm_segmentation_plot
```

# Evaluate true and false findings

The most important summaries in the script are the accuracy calculations. Since the last 500 features were assigned the condition effect, the script can treat `feature_id > 2500` as the truth set for differential abundance.

```{r accuracy-summaries}
res_labels <- tibble(
  isPos = c(TRUE, FALSE, TRUE, FALSE),
  truePos = c(TRUE, FALSE, FALSE, TRUE),
  label = c("True Positive", "True Negative", "False Positive", "False Negative")
)

accuracy_frame_ssc <- run_models_to_test_ssc_roi[[2]] |>
  mutate(
    isPos = p.adjust(p.value_condition1, method = "fdr") < 0.05,
    truePos = feature_id > 2500
  ) |>
  group_by(isPos, truePos) |>
  count() |>
  ungroup() |>
  full_join(res_labels, by = join_by(isPos, truePos)) |>
  mutate(n = replace_na(n, 0))

accuracy_frame_oracle <- run_models_to_test_oracle_roi[[2]] |>
  mutate(
    isPos = p.adjust(p.value_condition1, method = "fdr") < 0.05,
    truePos = feature_id > 2500
  ) |>
  group_by(isPos, truePos) |>
  count() |>
  ungroup() |>
  full_join(res_labels, by = join_by(isPos, truePos)) |>
  mutate(n = replace_na(n, 0))

accuracy_frame_skm <- run_models_to_test_skm_roi_7c[[2]] |>
  mutate(
    isPos = p.adjust(p.value_condition1, method = "fdr") < 0.05,
    truePos = feature_id > 2500
  ) |>
  group_by(isPos, truePos) |>
  count() |>
  ungroup() |>
  full_join(res_labels, by = join_by(isPos, truePos)) |>
  mutate(n = replace_na(n, 0))

accuracy_frame_ssc
accuracy_frame_oracle
accuracy_frame_skm
```

This is the payoff of the simulation. Because the truth is known, the downstream analysis can be evaluated in terms that are usually unavailable in real MSI data. We can ask not only whether a method finds signal, but whether it finds the correct signal and how much spurious discovery it produces.

# Volcano-style comparison across methods

The script also combines the model outputs into a shared plotting frame and colors findings by whether they are true positives, false positives, or non-significant.

```{r volcano-comparison, fig.width=10, fig.height=6}
bind_rows(
  run_models_to_test_oracle_roi[[2]] |>
    mutate(
      fdr.p.value_condition1 = p.adjust(p.value_condition1, method = "fdr"),
      isPosAdj = fdr.p.value_condition1 < 0.05,
      truePos = feature_id > 2500,
      id = "oracle"
    ),
  run_models_to_test_ssc_roi[[2]] |>
    mutate(
      fdr.p.value_condition1 = p.adjust(p.value_condition1, method = "fdr"),
      isPosAdj = fdr.p.value_condition1 < 0.05,
      truePos = feature_id > 2500,
      id = "ssc"
    ),
  run_models_to_test_skm_roi_7c[[2]] |>
    mutate(
      fdr.p.value_condition1 = p.adjust(p.value_condition1, method = "fdr"),
      isPosAdj = fdr.p.value_condition1 < 0.05,
      truePos = feature_id > 2500,
      id = "skm"
    )
) |>
  mutate(
    fdr_finding = ifelse(isPosAdj,
                         ifelse(truePos, "True Positive", "False Positive"),
                         ifelse(truePos, "False Negative", "True Negative")),
    fdr_finding_pos_only = factor(
      ifelse(isPosAdj,
             ifelse(truePos, "True Positive", "False Positive"),
             "True/False Negative"),
      levels = c("True Positive", "False Positive", "True/False Negative")
    ),
    id = factor(id, levels = c("oracle", "ssc", "skm"), labels = c("Segmentation with single\nknown informative marker", "Segmentation with Spatial\nShrunken Centroids", "Segmentation with Spatial K Means"))
  ) |>
  ggplot(aes(x = estimate_condition1, y = -log(fdr.p.value_condition1), color = fdr_finding_pos_only)) +
  geom_point(alpha = 0.65) +
  facet_grid(. ~ id) +
  scale_color_manual(
    values = c(
      "False Positive" = "#E1BE6A",
      "True Positive" = "#40B0A6",
      "True/False Negative" = "grey70"
    ),
    breaks = c("False Positive", "True Positive")
  ) +
  labs(
    x = "Estimated difference of simulated condition means in ROI",
    y = "-Log(FDR Adjusted P Value)",
    color = "Finding"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(file.path("Recreated-Figures","SI-Figure-Simualtion-1-ROI-Volcano-Plots.png"), height = 5, width = 8, units = "in")
```

This plot is especially useful because it separates two ideas that often get conflated: effect-size estimation and statistical significance. A segmentation method may recover a roughly correct ROI boundary yet still distort the estimated effect sizes or significance pattern.

# Feature-overlap diagnostics

The script concludes with Venn diagrams comparing top-ranked features from selected SSC and SKM segmentations to the true differentiating feature set. I view these as supporting diagnostics rather than the primary result, but they are still informative because they show whether the methods are prioritizing the same mass features as the truth set.

```{r overlap-diagnostics, fig.width=8, fig.height=8, results='hide', message=FALSE, warning=FALSE}
library(VennDiagram)
library(RColorBrewer)

ssc_control_charting_sample <- sscl[[which(samples$id == control_sample_id)]][[results_list[[1]][[which(results_list[[3]]$id == control_sample_id)]]$ssc_index]]
ssc_trt_charting_sample <- sscl[[which(samples$id == trt_sample_id)]][[results_list[[1]][[which(results_list[[3]]$id == trt_sample_id)]]$ssc_index]]

ssc_venn_args <- list(
  x = list(topFeatures(ssc_control_charting_sample, n = 500)$i,
           topFeatures(ssc_trt_charting_sample, n = 500)$i,
           2501:3000),
  category.names = c("Top 500 Features\nfrom a control (Cond B) image",
                     "Top 500 Features\nfrom a treatment (Cond A) image",
                     "True differentiating features"),
  disable.logging = TRUE,
  imagetype = "png",
  height = 1000,
  width = 1000,
  resolution = 300,
  compression = "lzw",
  pointsize = 18,
  lwd = 2,
  lty = "blank",
  fill = brewer.pal(3, "Pastel2"),
  euler.d = TRUE,
  sep.dist = 2,
  fontfamily = "sans",
  cat.cex = 0.6,
  cat.fontfamily = "sans",
  margin = 0.1
)

ssc_venn_file <- file.path("Recreated-Figures", "Figure-ROI-Simulation-1-SSC-Venn-Diff-Abun-Shared-Features.png")
do.call(
  venn.diagram,
  c(
    ssc_venn_args,
    list(
      filename = ssc_venn_file,
      output = TRUE
    )
  )
)
knitr::include_graphics(ssc_venn_file)


ssc_venn_roidiff_args <- list(
  x = list(topFeatures(ssc_control_charting_sample, n = 100)$i,
           topFeatures(ssc_trt_charting_sample, n = 100)$i,
           1:100),
  category.names = c("Top 100 Features\nfrom a control (Cond B) image",
                     "Top 100 Features\nfrom a treatment (Cond A) image",
                     "True 100 ROI-defining features"),
  disable.logging = TRUE,
  imagetype = "png",
  height = 1000,
  width = 1000,
  resolution = 300,
  compression = "lzw",
  pointsize = 18,
  lwd = 2,
  lty = "blank",
  fill = brewer.pal(3, "Pastel2"),
  euler.d = TRUE,
  sep.dist = 2,
  fontfamily = "sans",
  cat.cex = 0.6,
  cat.fontfamily = "sans",
  margin = 0.1
)

ssc_venn_roidiff_file <- file.path("Recreated-Figures", "Figure-ROI-Simulation-1-SSC-Venn-ROI-Def-Shared-Features.png")
do.call(
  venn.diagram,
  c(
    ssc_venn_roidiff_args,
    list(
      filename = ssc_venn_roidiff_file,
      output = TRUE
    )
  )
)
knitr::include_graphics(ssc_venn_roidiff_file)

skm_control_charting_sample <- skml_7c[[which(samples$id == control_sample_id)]]
skm_trt_charting_sample <- skml_7c[[which(samples$id == trt_sample_id)]]

skm_venn_args <- list(
  x = list(topFeatures(skm_control_charting_sample, n = 500)$i,
           topFeatures(skm_trt_charting_sample, n = 500)$i,
           2501:3000),
  category.names = c("Top 500 Features\nfrom a control (cond B) image",
                     "Top 500 Features\nfrom a treatment (cond A) image",
                     "True differentiating features"),
  disable.logging = TRUE,
  imagetype = "png",
  height = 1000,
  width = 1000,
  resolution = 300,
  compression = "lzw",
  pointsize = 18,
  lwd = 2,
  lty = "blank",
  fill = brewer.pal(3, "Pastel2"),
  euler.d = TRUE,
  sep.dist = 2,
  fontfamily = "sans",
  cat.cex = 0.6,
  cat.fontfamily = "sans",
  margin = 0.1
)

skm_venn_file <- file.path("Recreated-Figures", "Figure-ROI-Simulation-1-SKM-Venn-Diff-Abun-Shared-Features.png")
do.call(
  venn.diagram,
  c(
    skm_venn_args,
    list(
      filename = skm_venn_file,
      output = TRUE
    )
  )
)
knitr::include_graphics(skm_venn_file)



skm_venn_roidef_args <- list(
  x = list(topFeatures(skm_control_charting_sample, n = 100)$i,
           topFeatures(skm_trt_charting_sample, n = 100)$i,
           1:100),
  category.names = c("Top 100 Features\nfrom a control (cond B) image",
                     "Top 100 Features\nfrom a treatment (cond A) image",
                     "True differentiating features"),
  disable.logging = TRUE,
  imagetype = "png",
  height = 1000,
  width = 1000,
  resolution = 300,
  compression = "lzw",
  pointsize = 18,
  lwd = 2,
  lty = "blank",
  fill = brewer.pal(3, "Pastel2"),
  euler.d = TRUE,
  sep.dist = 2,
  fontfamily = "sans",
  cat.cex = 0.6,
  cat.fontfamily = "sans",
  margin = 0.1
)

skm_venn_roidef_file <- file.path("Recreated-Figures", "Figure-ROI-Simulation-1-SKM-Venn-ROI-Def-Shared-Features.png")
do.call(
  venn.diagram,
  c(
    skm_venn_roidef_args,
    list(
      filename = skm_venn_roidef_file,
      output = TRUE
    )
  )
)
knitr::include_graphics(skm_venn_roidef_file)
```

# Interpretation

This simulation is useful because it separates three pieces of the analysis that are intertwined in real data.

First, it defines a known spatial ROI. That gives us a benchmark for whether a segmentation method is recovering the intended region.

Second, it defines a known set of truly differential features. That gives us a benchmark for inference, including false discovery behavior.

Third, it passes the data through the same style of workflow used in the rest of the project: segmentation, ROI summarization, and feature-wise modeling. That makes the results more informative than a purely abstract simulation.

The main limitation is that the simulation still reflects the assumptions built into its construction. The ROI shape is regular and the effect sizes are prespecified. That is not a flaw so much as a reminder that this simulation is best interpreted as a targeted stress test of the workflow, not as a complete generative model of biological MSI data.
